---
title: "Data completeness state"
author: "Dan Weinberger"
date: "4/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ExcessILI)
library(cdcfluview)
library(reshape2)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(plotly)
library(MMWRweek)
library(readr)
library(rjson)
library(htmlTable)
library(RSocrata)
library(pdftools)
library(readr)
library(gsubfn)
library(INLA)
library (RCurl)
library(rjags)
library(HDInterval)
library(pbapply)
library(parallel)

```
## Backfilling
#NOTE THE DATA ARE MISSING FOR JULY-OCT 2019

```{r}
#Data from Andrew: NOTE: it looks like July-oct 2019 is missing from these data
wk11 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data11.csv')
wk11$max.date.report <- as.Date('2020-03-14')
wk11$report.date <- as.Date('2020-03-27')

wk12 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data12.csv')
wk12$max.date.report <- as.Date('2020-03-21')
wk12$report.date <- as.Date('2020-04-03')

wk13 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data13.csv')
wk13$max.date.report <- as.Date('2020-03-28')
wk13$report.date <- as.Date('2020-04-10')

wk15 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data15.csv')
wk15$max.date.report <- as.Date('2020-04-11')
wk15$report.date <- as.Date('2020-04-17')

wk16 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data16.csv')
wk16$max.date.report <- as.Date('2020-04-18')
wk16$report.date <- as.Date('2020-04-24')

wk17 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data17.csv')
wk17$max.date.report <- as.Date('2020-04-25')
wk17$report.date <- as.Date('2020-05-01')

wk18 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data18.csv')
wk18$max.date.report <- as.Date('2020-05-02')
wk18$report.date <- as.Date('2020-05-08')

wk19 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data19.csv')
wk19$max.date.report <- as.Date('2020-05-09')
wk19$report.date <- as.Date('2020-05-15')


wk20 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data20.csv')
wk20$max.date.report <- as.Date('2020-05-16')
wk20$report.date <- as.Date('2020-05-22')

wk21 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data21.csv')
wk21$max.date.report <- as.Date('2020-05-23')
wk21$report.date <- as.Date('2020-05-29')

wk22 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data22.csv')
wk22$max.date.report <- as.Date('2020-05-30')
wk22$report.date <- as.Date('2020-06-05')

wk23 <- 
  read.csv('./Archives_FluView_state_data/State_Custom_Data23.csv')
wk23$max.date.report <- as.Date('2020-06-06')
wk23$report.date <- as.Date('2020-06-12')
#Read in national data


provis.list <- lapply(c('01','02','03','04','05','06','07','08','09','10','11','12','13','14','15','16','17','18','19','20','21','22'),
                      function(x){
  d1 <- read.csv(paste0('./Data/provisional_pi/provisional', '2019-2020','_','week_',x,'.csv'))
  names(d1) <- toupper(names(d1))
  d1$week.death <- MMWRweek2Date(d1$YEAR, d1$WEEK) + days (6)
  d1$max.date.report <- max(d1$week.death)
 # d1$report.date <- d1$max.date.report + days(12)
  d1$report.date <- mmwr_week_to_date(2020,week= as.numeric(x))+ days(12)
  d1$SUB.AREA <- 'US'
  d1$epiyr <- d1$YEAR
  d1$epiyr[d1$WEEK<=26] <- d1$YEAR[d1$WEEK<=26] - 1
  return(d1)
    })
nat.reports <- do.call('rbind.data.frame',provis.list)

all.reports <- rbind.data.frame(wk11,wk12,wk13,wk15, wk16, wk17, wk18, wk19, wk20, wk21, wk22, wk23)


all.reports$epiyr <-
  as.numeric(as.character(substr(all.reports$SEASON,1,4)))

all.reports$year <- all.reports$epiyr
all.reports$year[all.reports$WEEK<=26] <-
  all.reports$epiyr[all.reports$WEEK<=26] +1
all.reports$week.death <-
  mmwr_week_to_date(all.reports$year, all.reports$WEEK)+6

#Fix formatting for the count variables
all.reports$NUM.INFLUENZA.DEATHS <-
  gsub(',','',all.reports$NUM.INFLUENZA.DEATHS) 
all.reports$NUM.INFLUENZA.DEATHS <-  as.numeric(as.character(all.reports$NUM.INFLUENZA.DEATHS)) 

all.reports$NUM.PNEUMONIA.DEATHS <-
  gsub(',','',all.reports$NUM.PNEUMONIA.DEATHS) 
all.reports$NUM.PNEUMONIA.DEATHS <-  as.numeric(as.character(all.reports$NUM.PNEUMONIA.DEATHS)) 

all.reports$TOTAL.DEATHS <-
  gsub(',','',all.reports$TOTAL.DEATHS) 
all.reports$TOTAL.DEATHS <-  as.numeric(as.character(all.reports$TOTAL.DEATHS))

nat.reports2 <- 
  nat.reports[, c('week.death','SUB.AREA','report.date', 'ALL.DEATHS')]
names(nat.reports2) <- c("week.death","SUB.AREA" ,"report.date","TOTAL.DEATHS")

all.reports2 <- rbind.data.frame(all.reports[,c("week.death","SUB.AREA" ,"report.date","TOTAL.DEATHS") ], nat.reports2)
```

```{r}
#Reporting delays from NobBS
delays <- readRDS('./outputs/NobBs.complete.iters.rds')
dimnames(delays)[[3]][dimnames(delays)[[3]]=='US'] <- 'United States'
#dimnames(delays)[[3]] <-
 # state.abb2[match(dimnames(delays)[[3]] , state.name2)]

#states1.match <- unique()
#states2.match <- dimnames(delays)[[3]]

#commn.states <- Reduce(intersect, list(states1.match,states2.match))

#delays <- delays[,,commn.states]
delays <- delays[,1:10000,]
delays.m <- melt(delays)
names(delays.m) <- c('time.since.death','iter','state','prop')
delays.med <- apply(delays,c (1,3),median)
delays.med.m <- melt(delays.med)
delays.med.m$Var1 <- as.numeric(as.character(delays.med.m$Var1))
names(delays.med.m) <- c('weeks.since.death','state','prop.complete')
```

```{r}
#Merge in reporting delay info
all.reports2$vintage <- all.reports2$report.date
all.reports2$week_end <- all.reports2$week.death 

all.reports2$n.weeks.ago <- round(as.numeric(difftime(all.reports2$vintage , all.reports2$week_end, units='weeks')))

#Merge in reporting delay
analysis.data <- merge(all.reports2,delays.med.m, by.x=c('SUB.AREA','n.weeks.ago'), by.y=c('state',"weeks.since.death"), all=T)

analysis.data <- 
  analysis.data[analysis.data$n.weeks.ago >= 2 ,]
```

##Estimate how many deaths will ultimately be reported
```{r}
analysis.data$estimated.deaths <-
  analysis.data$TOTAL.DEATHS / analysis.data$prop.complete

comp1 <- analysis.data[, c('SUB.AREA','week_end','estimated.deaths','n.weeks.ago')]
comp1.m <- melt(comp1, id.vars=c('SUB.AREA','week_end','n.weeks.ago'))

comp1.c <-dcast(comp1.m, SUB.AREA+week_end ~ n.weeks.ago)

comp1.c <- 
  comp1.c[comp1.c$week_end >= as.Date('2020-03-08'),]
```

#across all states, how do the estimates change 2,3,4... weeks out, over time?

Each line on this plot represents an estimate for that date 2,3,4.. weeks over time. This shows that the estimates from 2 weeks ago are an underestimate for recent dates, but they stabilize by 3 week
```{r, fig.width=5, fig.height=5}
comp1.c.agg <- aggregate(comp1.c[,-c(1:2)], by=list('week'=comp1.c$week_end), FUN=sum, na.rm=T)
comp1.c.agg[,-c(1)] <- apply(comp1.c.agg[,-c(1)],2, function(x){
  x[x==0] <- NA
  return(x)
  })
alpha.set=0.75
plot.cols <-
 c( rgb(228/256,26/256,28/256, alpha=alpha.set),
rgb(55/256,126/256,184/256, alpha=alpha.set),
rgb(77/256,175/256,74/256, alpha=alpha.set),
rgb(152/256,78/256,163/256, alpha=alpha.set),
rgb(55/256,127/256,0/256, alpha=alpha.set)
)

comp1.c.agg <- comp1.c.agg[,c(1:6)]
matplot(comp1.c.agg$week, comp1.c.agg[,-1], pch=c(16:20),bty='l', ylim=c(0,max(comp1.c.agg[,-1], na.rm=T) ), xaxt='n', col=plot.cols, ylab='Deaths adjusted for reporting delays', xlab='')
axis(side=1, at=as.Date(c('2020-03-01','2020-04-01','2020-05-01','2020-06-01')) , label=c('2020-03-01','2020-04-01','2020-05-01','2020-06-01'))
legend('bottomleft', legend=c('2 weeks', '3 weeks', '4 weeks', '5 weeks', '6 weeks'), ncol=2, pch=c(16:20),box.lty=0,inset=0.01, col=plot.cols)

```

## how much do thing increase N weeks out compared to 2 weeks, across all states
```{r}
comp1.c.agg.change <-t(apply(comp1.c.agg[,-1], 1, function(x){
  x<- x/x[1]
  return(x)
})
)
matplot(comp1.c.agg.change[,-1], type='l',xlab='Week of death', ylab='Relative change from deaths reported fater 2 weeks')
abline(h=1)

```

## look by state
There are a handful of states where the etsimates from 2 weeks ago are a severe underestimate, but stabilize after 3 weeks
```{r}
comp1.c.change <-t(apply(comp1.c[,-c(1:2)], 1, function(x){
  x<- x/x[1]
  return(x)
})
)
comp1.c.change <-
  cbind.data.frame(comp1.c[,c(1:2)],comp1.c.change)
matplot(comp1.c.change$week_end,
        comp1.c.change[,c(4:9)], type='p',xlab='Week of death', ylab='Relative change from deaths reported fater 2 weeks')
abline(h=1)
```


```{r}

```





