---
title: 'Estimating the early death toll of COVID-19 in the United States'
author: "Dan Weinberger"
date: "4/3/2020"
output:
  html_document:
    df_print: paged
    html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo=F,
  warning=FALSE, 
  message=FALSE,
  comment = "#>",
  dev=c('pdf', 'png'),
  fig.path='./figures/'
)
```


```{r, eval=F}
if (!require("devtools")) {
  install.packages("devtools")
}
devtools::install_github("weinbergerlab/ExcessILI")
```

```{r setup}
library(ExcessILI)
library(cdcfluview)
library(reshape2)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(plotly)
library(MMWRweek)
library(readr)
library(rjson)
library(htmlTable)
library(RSocrata)
source('./functions/ts_plot_func.R')
```


```{r}
plot.states <- c('CA','TX','FL', 'NY','PA','IL','OH','GA','MI','WA' )
```


```{r}
#cross walk file to map  states to hhs regions
# 
hhs_states <- cdcfluview::hhs_regions
hhs_states$state <- state.abb[match(hhs_states$state_or_territory, state.name)]
hhs_states$state_region <- hhs_states$region
hhs_states$state_region[hhs_states$state %in% plot.states] <- hhs_states$state[hhs_states$state %in% plot.states]
hhs_states <- hhs_states[,c('state','state_region')]
hhs_states$state_region <- gsub(' ','', hhs_states$state_region)
hhs_states.cw <- hhs_states[!is.na(hhs_states$state),]

hhs_states.cw.spl <- split(hhs_states.cw, hhs_states.cw$state_region)

hhs_states.cw.spl <-lapply(hhs_states.cw.spl, function(x) {
  x$state_region.lab <- paste(x$state, collapse=',')
  return(x)
  }
)

hhs_states.cw <-do.call('rbind.data.frame',hhs_states.cw.spl)
hhs_states.cw$state_region[hhs_states.cw$state_region != hhs_states.cw$state] <- hhs_states.cw$state_region.lab[hhs_states.cw$state_region != hhs_states.cw$state]
```

```{r}
#TESTING DATA FORMATTING
url.test<-"https://covidtracking.com/api/v1/states/daily.json" 


json_data <- fromJSON(file=url.test)
test.dates <- as.character(sapply(json_data,'[[','date'))

test.state <- sapply(json_data,'[[','state')
testN <- sapply(json_data, '[[', 'totalTestResultsIncrease')
testN <-sapply(testN, function(x){ 
  if(is.null(x)){
  x <-0
  }
  return(x)
  }
  )
deathN <- sapply(json_data, '[[', 'deathIncrease')
deathN <-sapply(deathN, function(x){ 
  if(is.null(x)){
  x <-0
  }
  return(x)
  }
  )
test.ds <- cbind.data.frame('state'=test.state,test.dates, 'testN.day'=testN, 'deathN.day'=deathN )
test.ds$test.date.wk <-floor_date(as.Date(test.ds$test.dates, '%Y%m%d'),'week')

test.ds <- merge(test.ds, hhs_states.cw, by='state')


test.ds.agg <- aggregate( test.ds[,c('testN.day','deathN.day')] , by=list('state'=test.ds$state_region,'date'=test.ds$test.date.wk), FUN=sum)
names(test.ds.agg) <- c('state','date','testN','deathN')

pop1<-read.csv('./Data/nst-est2019-01.csv')
pop1$state_name <- substring(pop1$state_name,2)
pop1$state <- state.abb[match(pop1$state_name, state.name)]
pop1$census_bureau_pop_2019 <- gsub( ',', '',pop1$census_bureau_pop_2019)
pop1$census_bureau_pop_2019 <-as.numeric(pop1$census_bureau_pop_2019)

pop1.reg <- merge(pop1, hhs_states.cw, by='state')
pop2 <- aggregate(pop1.reg[,"census_bureau_pop_2019"], by=list('state'=pop1.reg$state_region), FUN=sum)
names(pop2) <- c('state','census_bureau_pop_2019')

test.ds2 <- merge(test.ds.agg, pop2, by='state')
test.ds2$test.week.per.capita <- test.ds2$testN/test.ds2$census_bureau_pop_2019*1000
names(test.ds2) <- c('state','date','testN','covid.track.death', 'pop2019',"test.week.per.capita")

test.ds3.spl <- test.ds2
```

```{r}
#download the NREVSS data
nrevvs.state <- cdcfluview::who_nrevss(region = c("state"))
  
  clin <- nrevvs.state[["clinical_labs"]]
  clin$state <- state.abb[match(clin$region, state.name)]
  
  data(cdcfluview::hhs_regions)
  
  cw.file <- cdcfluview::hhs_regions
  
  clin2 <- merge(clin, hhs_states.cw,
                 by = "state")

  clin2 <- merge(clin2, cw.file,
                 by.x = "region",
                 by.y = "state_or_territory")
  
  clin2.subsetvars <- 
    c('region', 'region_number',
      'year', 'week', 'wk_date',
      'total_a','total_b',
      'total_specimens','state_region')
  
  clin2 <- clin2[, clin2.subsetvars]
  
  names(clin2)[1:2] <- c("state", "hhs_region")
  
  clin2$total_specimens <- as.numeric(clin2$total_specimens)
  clin2$total_a <- as.numeric(clin2$total_a)
  clin2$total_b <- as.numeric(clin2$total_b)
  
  clin2 <- aggregate(clin2[, c('total_specimens','total_a','total_b')], by=list('state'=clin2$state_region, 'wk_date'=clin2$wk_date,'hhs_region'=clin2$hhs_region), FUN=sum, na.rm=T)
  
  ##Florida doesn't have ILI data, so use regions ILI dat
  nrevvs_hhs <- cdcfluview::who_nrevss(region = c("hhs"))
  
  clin.hhs <- nrevvs_hhs[["clinical_labs"]]
  clin.hhs.subsetvars <-
    c('region',
      'wk_date',
      "total_a",'total_b',
      'total_specimens')
  
  clin.hhs <- clin.hhs[, clin.hhs.subsetvars]
  clin.hhs$region <- as.numeric(gsub("Region ", "", clin.hhs$region))
  
  names(clin.hhs) <-
    c("hhs_region",
      "wk_date",
      "hhs_total_a",'hhs_total_b',
      'hhs_total_specimens')
  
  clin3 <- merge(clin2, clin.hhs,
                 by = c("hhs_region", "wk_date"))

  #If no specimens for state, use HHS region estimates
  clin3$total_specimens[clin3$total_specimens==0] <-
    clin3$hhs_total_specimens[clin3$total_specimens==0]
  
  clin3$total_a[clin3$total_specimens==0] <-
    clin3$hhs_total_a[clin3$total_specimens==0]
  
  clin3$total_b[clin3$total_specimens==0] <-
    clin3$hhs_total_b[clin3$total_specimens==0]
  
  clin3$total_specimens <- as.numeric(clin3$total_specimens)
  clin3$total_a <- as.numeric(clin3$total_a)
  clin3$total_b <- as.numeric(clin3$total_b)  
  
  clin3$flu_pct_adj <- (clin3$total_a + clin3$total_b)/clin3$total_specimens
  
   clin4<-clin3[,c('state','flu_pct_adj', 'wk_date')]

  clin4.lag1<-clin4
  clin4.lag1$wk_date <- clin4$wk_date + days(7)
  names(clin4.lag1) <-c('state','flu_pct_adj_lag1','wk_date')
  
  clin4.lag2<-clin4
  clin4.lag2$wk_date <- clin4$wk_date + days(14)
   names(clin4.lag2) <-c('state','flu_pct_adj_lag2','wk_date')
   
clin4.lags <- merge(clin4, clin4.lag1, by=c('state','wk_date'))
clin4.lags <- merge(clin4.lags, clin4.lag2, by=c('state','wk_date'))


```

```{r, include=F}
#Format and fill mssings with 0s

## Download the mortality data
#pi.data.alt <- read.socrata("https://data.cdc.gov/resource/pp7x-dyj2.json")
#pi.data.alt <- pi.data.alt[pi.data.alt$geoid=='state',]
pi.data <- pi_mortality(coverage_area='state')
pi.data$percent_complete[pi.data$percent_complete > 1] <- 1

pi.data$state <- state.abb[match(pi.data$region_name, state.name)]
pi.data$state[pi.data$region_name == 'New York City'] <- 'NYC'

pi.data <- merge(pi.data,hhs_states.cw, by='state', all=T)
pi.data$state_region[pi.data$state=='NYC'] <- 'NYC'
pi.data$pct.comp.wgt <- pi.data$percent_complete*pi.data$all_deaths

pi.data.agg <- aggregate(pi.data[,c('total_pni','all_deaths','pct.comp.wgt')], by=list('state'=pi.data$state_region, 'week_start'=pi.data$week_start, 'week_end'=pi.data$week_end), FUN=sum, na.rm=T)
pi.data.agg$percent_complete <- pi.data.agg$pct.comp.wgt/pi.data.agg$all_deaths

spl1<-split(pi.data.agg, pi.data.agg$state)
min.state <- lapply(spl1, function(x){ x$miss.x<-min(x$total_pni, na.rm=T)
return(x)
                })
pi.data.clean <- do.call('rbind.data.frame',min.state)
pi.data.clean <- pi.data.clean[!is.na(pi.data.clean$miss.x),]

pi.data.clean2<- merge( pi.data.clean,clin4.lags, by.x=c('week_start', 'state'), by.y=c('wk_date','state'))

pi.data.clean2<- pi.data.clean2[order(pi.data.clean2$state, pi.data.clean2$week_start),]
states.cdc <- unique(pi.data.clean2$state)

date.print <- max(pi.data.clean2$week_end)

```

```{r, eval=F}
#What is most appropriate lag to use for NREVSS data?
pi.data.clean2.spl <- split(pi.data.clean2, pi.data.clean2$state)
cor.lags <- sapply(pi.data.clean2.spl, function(x){
   cor(x[,c("percent_pni","flu_pct_adj",'flu_pct_adj_lag1','flu_pct_adj_lag2')])
 }, simplify='array')
matplot(cor.lags['percent_pni',-1,], type='l')

#Is lag of 0,1,or 2 the best?...shows 1 weeks lag best in 20 states,  2 week lag best in 3 states
table(apply(cor.lags['percent_pni',-1,],2, function(x) which(max(x)==x)))

```

## Abstract
Mortality data on deaths due to pneumonia & influenza through the week ending `r  date.print`

*Background*
Tracking the severity and public health impact of emerging diseases, such as the novel coronavirus diseases COVID-19, is a critical need. These efforts are often hampered by testing issues and reporting lags for key epidemiological data. Evaluating unexplained increases in deaths attributed to non-specific causes, such as pneumonia, can provide a more complete picture of the burden caused by COVID-19. 

*Methods*
We evaluated increases in the occurrence of deaths due to P&I above a seasonal baseline across the United States and compared the rate of excess deaths to reported rates of death due to COVID-19, as well as state-level coronavirus testing rates and excess rates of influenza-like illness. 

*Results*
There were notable increases in the rate of death due to pneumonia and influenza. In a number of states, these increases pre-dated the increase in testing rates and went uncounted. There was substantial variability between states in the discrepancy between reported rates of death due to COVID-19 and the estimated burden of excess deaths due to P&I. In some states, the burden of reported deaths was 10-fold lower than the burden of excess deaths due to P&I. 

*Conclusions*
Given the lack of consistent virological testing, tracking spikes in deaths due to non-specific causes provides the most complete estimate of the burden of COVID-19. This approach provides a framework for tracking the progression of the epidemic and its public health impact. 

```{r, include=F}
#Run analysis
excess_deaths1.adjusted <-
  excessCases(ds = pi.data.clean2,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "flu_pct_adj_lag1",
              #covs=c("flu_pct_adj", "flu_pct_adj_lag1", "flu_pct_adj_lag2"),
              use.syndromes = c("total_pni"),
              extrapolation.date = "2020-03-01",
              time.res='week')
```


```{r, echo=F}
 #dashboardPlot(excess_deaths1.adjusted)
```


```{r}
### Extract the quantities of interest

#Which syndrome do you want to plot, and over what time range?

syndrome.select <- 'total_pni' #which syndrome do you want to plot?
n.days<-52  #How many days to plot?
ds <- excess_deaths1.adjusted
```
 
```{r}
#Extract the data needed to plot from the results

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred<-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr <-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

excess_deaths <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

excess_deaths_var <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred.var")
```

## Figure 1: Observed weekly death rate vs seasonal baseline (+/-95% Prediction Interval) 
 The black line shows the observed proportion of deaths that were due to Pneumonia & Influenza (P&I) per week. The red line and shaded area represent the 95% Prediction Interval. The latest P&I data is for the week ending `r  max(dates1)+7`.

```{r }
n.days <- 52
select.indices <- (length(dates1)-n.days):length(dates1)
dates<-dates1[select.indices]
states <- dimnames(pred)[[2]]
ages <- dimnames(pred)[[3]]
```

```{r fig1, fig.width=7.3, fig.height=6.9}

  par(mfrow=c(5,4), mar=c(2,4,2,1))
states.cdc.order <- c(states.cdc[states.cdc %in% state.abb],
                      states.cdc[!(states.cdc %in% state.abb)]) 
plot.state.indices <- match( states.cdc.order,dimnames(pred)[[2]]  )
for(i in plot.state.indices){
  for(j in 1:dim(pred)[3]){
    y.range1<-range(c( pred[select.indices,i,j]/denom[select.indices,i,j],obs[select.indices,i,j]/denom[select.indices,i,j], upi[select.indices,i,j]/denom[select.indices,i,j],0))
    if(states[i] %in% state.abb ){
      state.name.plot <-    
      state.name[match(states[i],state.abb)]
    }else{
    state.name.plot <- states[i]
  }
  plot(dates,
       pred[select.indices,i,j]/denom[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       xlab='',
       ylab='Proportion',
       #main=paste(states[i])
       )
  
  text(x=as.Date('2019-03-01'),y=0.005, state.name.plot, pos=4,cex=0.9)

  points(dates,
         obs[select.indices,i,j]/denom[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates,
            rev(dates)),
          c(lpi[select.indices,i,j]/denom[select.indices,i,j],
            rev(upi[select.indices,i,j]/denom[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

## Figure 2: Observed deaths/expected deaths by state
Excess Deaths due to pneumonia and influenza

```{r, echo=F}
rr2<-rr[,,1]
date.mmwrdates <- mmwr_week(dates1)
mmwr.epiyr<- date.mmwrdates$mmwr_year
mmwr.epiyr[date.mmwrdates$mmwr_week<=26] <- mmwr.epiyr[date.mmwrdates$mmwr_week<=26] - 1

mmwr.epiwk <- date.mmwrdates$mmwr_week
mmwr.epiwk[date.mmwrdates$mmwr_week>=27]<-date.mmwrdates$mmwr_week[date.mmwrdates$mmwr_week>=27] - 52
mmwr.epiwk <- mmwr.epiwk +26
check<-cbind.data.frame(date.mmwrdates,mmwr.epiwk, mmwr.epiyr)
```

These plots show the Observed/Expected number of deaths due to pneumonia and influenza in each week for the 2019-20 year (red) compared to previous years (gray). Values close to 1 indicate that the values for that week are close to what would be expected based on the time of year and influenza activity.

```{r fig2, fig.width=7.3, fig.height=6.9}

  par(mfrow=c(5,4), mar=c(2,4,2,1))
states.cdc.order <- c(states.cdc[states.cdc %in% state.abb],
                      states.cdc[!(states.cdc %in% state.abb)]) 
plot.state.indices <- match( states.cdc.order,dimnames(pred)[[2]]  )

#plot.state.indices <- which(dimnames(pred)[[2]] %in% plot.states )

for(i in plot.state.indices){
    y.range1<-c(0.5,1.5)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    
    if(states[i] %in% state.abb ){
      state.name.plot <-    
      state.name[match(states[i],state.abb)]
    }else{
    state.name.plot <- states[i]
    }
    
    matplot(ds2.c$epiwk         ,
       exp(ds2.c[,-1]),
       type='l',
       col=cols1,
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected')
    abline(h=1, col='black')
     text(x=ds2.c$epiwk[1]*1.0 , y=y.range1[1]*1.08 , state.name.plot, pos=4, cex=0.9)
    }
```



```{r, echo=F, eval=F}
## excess deaths

excess_deaths2 <- excess_deaths[dates1 >= as.Date('2020-02-02'),,1]
excess_deaths.state <- apply(excess_deaths2,2,sum)
cumsum_excess_deaths_state <- apply(excess_deaths2,2,cumsum)
#matplot(cumsum_excess_deaths_state, type='l', bty='l')

```

## Figure 3: Reported number of COVID-19 deaths,compared with the excess deaths due to pneumonia and influenza in each week, by state. 
The blue line shows the number of excess P&I cases, the red solid line shows the reported number of COVID-19 deaths for the same week (as compiled by the Johns Hopkins CSEE), and the dotted red line shows the reported COVID-19 deaths for weeks in which the CDC data were not yet available. Excess P&I estimates from recent weeks are adjusted for incomplete records using the "percent_complete" variable.

```{r}
jh1.url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
jh1<-as.data.frame(read_csv(url(jh1.url)))
jh1 <- jh1[jh1$iso2=='US',]
jh1.m <- melt(jh1[, c(6:7,13:ncol(jh1))], id.vars=c('Admin2',"Province_State"))
jh1.m$variable <-as.Date(jh1.m$variable, '%m/%d/%y')

#Aggregate across the admin2 regions by day
jh2 <- aggregate( jh1.m$value, by=list('date'=jh1.m$variable, 'state'=jh1.m$Province_State), FUN=sum, na.rm=T)
names(jh2) <- c('date','state', 'cum.deaths')

jh2.spl <- split(jh2, jh2$state)
jh2.spl <-lapply(jh2.spl, function(x){
  x$deaths<-x$cum.deaths
  for(i in 2:nrow(x)){
    x$deaths[i] <- x$cum.deaths[i] - x$cum.deaths[(i-1)]
  }
  return(x)
})
jh2 <- do.call('rbind.data.frame', jh2.spl)
jh2$state <- state.abb[match(jh2$state, state.name)]

#round date to sunday
jh2$date <-floor_date(jh2$date, 'week')

jh2 <- merge(jh2, hhs_states.cw, by='state')

jh2.agg<- aggregate( jh2$deaths, by=list('state'=jh2$state_region, 'date'=jh2$date), FUN=sum)
names(jh2.agg) <- c('state', 'date', 'deaths')

excess_deaths.df <- cbind.data.frame('date'=dates1,excess_deaths[,,1])
excess_deaths.m <- melt(excess_deaths.df, id.vars =c('date'))
names(excess_deaths.m) <-c('date','state','excess_pi')

#excess_deaths_var
excess_deaths.var.df <- cbind.data.frame('date'=dates1,excess_deaths_var[,,1])
excess_deaths.var.m <- melt(excess_deaths.var.df, id.vars =c('date'))
names(excess_deaths.var.m) <-c('date','state','excess_pi_var')

excess_deaths.upi.df <- cbind.data.frame('date'=dates1,  (obs[,,1]-lpi[,,1] ))
excess_deaths.upi.m <- melt(excess_deaths.upi.df, id.vars =c('date'))
names(excess_deaths.upi.m) <-c('date','state','excess_deaths.upi')

excess_deaths.lpi.df <- cbind.data.frame('date'=dates1,(obs[,,1]-upi[,,1]) )
excess_deaths.lpi.m <- melt(excess_deaths.lpi.df, id.vars =c('date'))
names(excess_deaths.lpi.m) <-c('date','state','excess_deaths.lpi')

total_pi_df <- cbind.data.frame('date'=dates1,(obs[,,1]) )
total_pi_df.m <- melt(total_pi_df, id.vars =c('date'))
names(total_pi_df.m) <-c('date','state','total_pi')

#states.cdc <- unique(excess_deaths.m$state)
jh3 <- merge(jh2.agg, excess_deaths.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.lpi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.upi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, total_pi_df.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.var.m, by=c('state','date') , all=T)

#jh3 <-jh3[jh3$state %in% states.cdc,]


jh3 <- merge( jh3, pi.data.clean2[c('state', "week_start",'percent_complete')], by.x=c('state','date') , by.y=c('state',"week_start"), all=T)

jh3 <- jh3[jh3$date >= as.Date('2020-02-09'),] #2 weeks before 1st reported deaths by CDC
jh3$deaths[is.na(jh3$deaths)] <- 0

jh3 <- merge(jh3, test.ds3.spl, by=c('state','date'), all=T)
jh3$covid.track.death[is.na(jh3$covid.track.death)] <- 0

jh3 <- jh3[ !is.na(jh3$deaths), ]

#Only plot out to 1 week ahead of the US P&I data
jh3 <- jh3[ jh3$date <= max(pi.data$"week_start" +7),]
jh3 <- jh3[jh3$state %in% states.cdc,]
#jh3 <-jh3[!is.na(jh3$excess_pi ),]
```

```{r fig3, fig.width=7.3, fig.height=6.9}
## uses covidtracking.com data for reported deaths

par(mfrow=c(5,4), mar=c(2,3,1,3))
ts.plot.func(ds.plot=jh3, death.var='covid.track.death' ,states.plot=unique(jh3$state) )


```

## Table 1: Excess P&I deaths and reported deaths due to COVID-19

```{r, table1}
count.excess <-jh3[jh3$date >= as.Date('2020-02-09') & jh3$date <=max(pi.data$week_end) ,]

count.excess$excess_pi_adj <-count.excess$excess_pi/count.excess$percent_complete

count.excess$total_pi_adj <-count.excess$total_pi/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi_adj','total_pi_adj','covid.track.death','excess_pi_var')], by=list('state'=count.excess$state) ,FUN=sum)

count.excess2 <- merge(count.excess2, pop2, by='state')

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_ucl_INC <- (count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var))/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_ci_lcl_INC <- (count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var))/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_pi_INC <- count.excess2$excess_pi_adj/count.excess2$census_bureau_pop_2019*1000000

count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi_adj), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 
count.excess2$excess_pi_ci_INC <- paste0(round(count.excess2$exces_pi_INC,1), ' (',  round(count.excess2$exces_ci_lcl_INC,1), ', ', round(count.excess2$exces_ci_ucl_INC,1),')'  ) 

#sort by incidence
count.excess2 <- count.excess2[rev(order(count.excess2$exces_pi_INC)),]

count.excess2$total_pi_adj <- round(count.excess2$total_pi_adj)
count.excess3 <- count.excess2[, c('state','total_pi_adj','excess_pi_ci' ,'covid.track.death','excess_pi_ci_INC')]
names(count.excess3) <-c('State','Total P&I deaths', 'Excess P&I deaths', 'Reported COVID-19 deaths','Excess P&I deaths/1,000,000 people')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")

htmlTable(count.excess3, caption=paste0('Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from February 9, 2020 through ', last.date.format), rnames=F)
```


```{r}
## And finally,  use the NCHS COVID death data
#(Not yet available with state-level resolution)
#cdc.covid.url<- "https://data.cdc.gov/resource/hc4f-j6nb.json"
```


## Figure 4: Compare Excess P&I mortality vs Excess ILI

Here we compare the observed vs expected number of deaths due to pneumonia and influenza in each week compare to the observed vs expected number of outpatient visits for influenza-like illness (ILI) in each week. we would expect ILI (blue line) to increase earlier than deaths (red line)

```{r, include=F}
ili.data <- ilinet(region = c("state"))
ili.data$state <- state.abb[match(ili.data$region, state.name)]
ili.data       <- ili.data[, c("state", "week_start", "ilitotal", "total_patients")]
ili.data       <- ili.data[!is.na(ili.data$total_patients),]
ili.data.spl   <- split(ili.data, ili.data$state)

min<-sapply(ili.data.spl, function(x)  min(x$total_patients))

state.select<-names(min)[which(min>0) ]
ili.data <- ili.data[ili.data$state %in% state.select,]
## Run the main analysis function, adjusting for flu using NREVSS data

ili.data2 <- merge(ili.data, hhs_states.cw, by='state')
ili.data2.agg <- aggregate(ili.data2[,c('ilitotal','total_patients')], by=list('state'=ili.data2$state_region,"week_start"=ili.data2$week_start), FUN=sum)

ili.data2.agg <- merge(ili.data2.agg, clin4.lags, by.x=c('state','week_start'), by.y=c('state', 'wk_date'))

excess_cases1 <-
  excessCases(ds = ili.data2.agg,
              datevar       = "week_start", 
              statevar      = "state",
              denom.var     = "total_patients",
              adj.flu       = "flu_pct_adj",
              use.syndromes = c("ilitotal"),
              extrapolation.date = "2020-03-01",
              time.res='week')

dates.ili <-
  excess_cases1[[1]][[1]][[1]]$date
  
rr.ili <-  excessExtract(ds = excess_cases1,
                syndrome = "ilitotal",
                extract.quantity = "resid1")
rr2.ili <- rr.ili[,,1]

date.mmwrdates.ili <- mmwr_week(dates.ili)
mmwr.epiyr.ili<- date.mmwrdates.ili$mmwr_year
mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] <- mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] - 1

mmwr.epiwk.ili <- date.mmwrdates.ili$mmwr_week
mmwr.epiwk.ili[date.mmwrdates.ili$mmwr_week>=27]<-date.mmwrdates.ili$mmwr_week[date.mmwrdates.ili$mmwr_week>=27] - 52
mmwr.epiwk.ili <- mmwr.epiwk.ili +26
```

```{r fig4, fig.width=7.3, fig.height=8}
common.states <- intersect(colnames(rr2), colnames(rr2.ili) )

states.cdc.order <- c(common.states[common.states %in% state.abb],
                      common.states[!(common.states %in% state.abb)]) 
plot.state.indices <- match( states.cdc.order,common.states  )

rr2.comp <- rr2[,states.cdc.order]
rr2.ili.comp <- rr2.ili[,states.cdc.order] 
date.start <-unique(pi.data$week_start[pi.data$weeknumber==27 & year(pi.data$week_start)==2019])
dates.plot4 <- seq.Date(from=date.start, length.out=52, by='week')

  par(mfrow=c(5,4), mar=c(2,4,2,2))

for(i in 1:ncol(rr2.comp)){
    y.range1<-c(0,5)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2.comp[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    
    if(states.cdc.order[i] %in% state.abb ){
      state.name.plot <-    
      state.name[match(states.cdc.order[i],state.abb)]
    }else{
    state.name.plot <- states.cdc.order[i]
    }
    
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    plot(dates.plot4         ,
       exp(ds2.c[,'2019']),
       type='l',
       col='#e41a1c',
       ylim=y.range1,
       xlim=c(as.Date(c('2019-07-01')),max(pi.data$week_end)),
       bty='l',
       lty=1,
       xlab='',
       xaxt='n',
       ylab='Observed/Expected'
       #main=common.states[i]
       )
    axis(1, at=as.Date(c('2019-07-01', '2019-10-01','2020-01-01', '2020-04-01', '2020-07-01')) , labels= c('Jul','Oct','Jan','Apr','Jul')  )

  es2<-cbind.data.frame('epiwk'=mmwr.epiwk.ili,'epiyr'=mmwr.epiyr.ili, rr=rr2.ili.comp[,i])
    es2.c<-dcast(es2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
     points(dates.plot4         ,
       exp(es2.c[,'2019']), type='l', col='#377eb8')
    abline(h=1, col='gray', lty=2)
    
    text(min(dates.plot4), y=0.2, state.name.plot,pos=4, cex=0.9)
}

```

## Supplementary figure: Excess all-cause deaths

Now instead of modeling the proportion of deaths that were due to pneumonia and influenza, we will model all-caused deaths
```{r fig.s1}
#Run analysis
pi.data.clean2$one <- 1
pi.data.clean2$percent_complete.adj <-pi.data.clean2$percent_complete - 0.5
excess_deaths1.ac <-
  excessCases(ds = pi.data.clean2,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "percent_complete.adj",
              adj.flu       = "flu_pct_adj_lag1",
              #covs=c("flu_pct_adj", "flu_pct_adj_lag1", "flu_pct_adj_lag2"),
              use.syndromes = c("all_deaths"),
              extrapolation.date = "2020-03-01",
              time.res='week')
```

```{r}
ds <- excess_deaths1.ac
syndrome.select <- 'all_deaths'
excess_deaths.ac <- excessExtract(ds = excess_deaths1.ac,
                syndrome = 'all_deaths',
                extract.quantity = "unexplained.cases")
dates1.ac <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred.ac<-  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr.ac <-  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

```

```{r}
# compare excess all-cause deaths with excess P&I deaths

dates.select <- which(dates1 >= as.Date('2020-01-01'))
#plot( excess_deaths[dates.select,,1],excess_deaths.ac[dates.select,,1])

```

The black line shows the observed number of *all* deaths per week, regardless of cause. The red line and shaded area represent the 95% Prediction Interval. The latest data is for the week ending `r  max(dates1)+7`. Note that these are **adjusted** for percent completeness of the data. There are no obvious jumps in all-cause mortality, even in Washington State

```{r fig.s2, fig.width=8, fig.height=7}

  par(mfrow=c(4,4))

for(i in 1:dim(pred.ac)[2]){
  for(j in 1:dim(pred.ac)[3]){
    y.range1<-range(c( pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j], upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],0))
  plot(dates,
       pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       ylab='Proportion',
       main=paste(states[i],ages[j]))

  points(dates,
         obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates,
            rev(dates)),
          c(lpi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
            rev(upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

## Supp Figure: Extract official data directly from FL 
This shows that the Johns Hopkins data  are off by ~ 1 week in the date assignments compared to Florida official stats. The Florida data records the 'date case counted', so it is possible this date is not date of death but rather the date when the case was recognized.

```{r fig.s3}
library(pdftools)
library(readr)
library(gsubfn)
fl_text <-pdf_text("./Data/covid-19-data---daily-report-2020-04-06-1657.pdf")
fl1<-read_lines(fl_text)
death.start <- grep('Death    County         Age Gender', fl1 )
death.start.gap <-death.start +3

fl2 <-fl1[death.start.gap[1] : (death.start.gap[1]+300)]
fl.dates <- strapplyc(fl2, "[0-9/]{8,}", simplify = TRUE)
fl.dates <- unlist(fl.dates)
fl.dates <- as.Date(fl.dates, '%m/%d/%y')
fl3 <-cbind.data.frame('date'=fl.dates,'deaths'=rep(1, length(fl.dates)))
fl3$wkdate <- floor_date(fl3$date, 'week')
fl4 <- aggregate(fl3[,'deaths'], by=list('date'=fl3$wkdate), FUN=sum )
names(fl4) <- c('date', 'official_deaths')

fl5 <- merge(jh3, fl4, by='date', all=T)
fl5 <-fl5[fl5$state=='FL',]
fl5 <-fl5[!is.na(fl5$state),]

par(mfrow=c(1,1))
states.cdc.fl <-'FL'
y2.range.test <- range(fl5$test.week.per.capita, na.rm=T)
for(i in 1:length(states.cdc.fl)){
    ds.select <- fl5[fl5$state==states.cdc.fl[i],]
   # y.range1<-range(c(ds.select$covid.track.death, ds.select$excess_pi/ds.select$percent_complete), na.rm=T)
    
     ave_pi <- mean(ds.select$total_pi, na.rm=T)
  ave.range <- c(-ave_pi*0.3, ave_pi*0.7)
  y.range1<-range(c(ds.select[,death.var], ds.select$excess_pi/ds.select$percent_complete), na.rm=T)
 
    ds.select$death.early <- ds.select$covid.track.death
    ds.select$death.early[is.na(ds.select$excess_pi)] <- NA
    par(new=FALSE)
    plot(ds.select$date         ,
       ds.select$covid.track.death,
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       lty=3,
       ylab='N Deaths',
       xlab='Date',
       main=paste(states.cdc.fl[i] ))
    
       points(ds.select$date         ,
       ds.select$death.early, type='l', col='red',
       lty=1, lwd=2)
       
      points(ds.select$date         ,
       ds.select$deaths, type='l', col='purple',#JH data
       lty=1, lwd=2)
       
      points(ds.select$date         ,
       ds.select$official_deaths, type='l', col='orange',
       lty=1, lwd=2)

       points(ds.select$date         ,
       ds.select$excess_pi/ds.select$percent_complete, type='l', col='blue', lwd=2)
    abline(h=1, col='black')
    
    par(new=TRUE)
   plot(ds.select$date, ds.select$test.week.per.capita, ylim=y2.range.test, type='l', lty=2, lwd=0.5, col='gray', yaxt='n',xaxt='n', ylab='', xlab='', bty='l')
   axis(side=4,at=c(0,1,2,3,4,5) , labels=F)
 mtext("Tests per 1000", side=4, line=0.2, col='gray', cex=0.75)
    
}

```

##Sensitivity analysis 1: Do not adjust for underreporting of deaths

```{r, tables1}
count.excess <-jh3[jh3$date >= as.Date('2020-02-09') & jh3$date <=max(pi.data$week_end) ,]

count.excess$excess_pi_adj <-count.excess$excess_pi/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi','excess_pi_adj','covid.track.death','excess_pi_var')], by=list('state'=count.excess$state) ,FUN=sum)

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi + 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi - 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_ucl_adj <- round(count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_lcl_adj <- round(count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var))


count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 

count.excess2$excess_pi_ci_adj <- paste0(round(count.excess2$excess_pi_adj ), '(',  count.excess2$exces_ci_lcl_adj , ', ', count.excess2$exces_ci_ucl_adj ,')'  ) 

count.excess3 <- count.excess2[, c('state','excess_pi_ci', 'excess_pi_ci_adj', 'covid.track.death')]
names(count.excess3) <-c('State', 'Excess P&I deaths (unadjusted)','Excess P&I deaths (adjusted)', 'Reported COVID-19 deaths')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")

htmlTable(count.excess3, caption=paste0('Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from February 9, 2020 through ', last.date.format), rnames=F)
```
```

## Sensitivity analysis 2: do not adjust for flu
```{r, include=F}
#Run analysis
excess_deaths1.unadjusted <-
  excessCases(ds = pi.data.clean2,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "none",
              use.syndromes = c("total_pni"),
              extrapolation.date = "2020-03-01",
              time.res='week')

syndrome.select <- 'total_pni' #which syndrome do you want to plot?
n.days<-52  #How many days to plot?
ds <- excess_deaths1.unadjusted
```
 
```{r}
#Extract the data needed to plot from the results

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred<-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr <-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

excess_deaths <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

excess_deaths_var <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred.var")
```

```{r}
jh1.m <- melt(jh1[, c(6:7,13:ncol(jh1))], id.vars=c('Admin2',"Province_State"))
jh1.m$variable <-as.Date(jh1.m$variable, '%m/%d/%y')

#Aggregate across the admin2 regions by day
jh2 <- aggregate( jh1.m$value, by=list('date'=jh1.m$variable, 'state'=jh1.m$Province_State), FUN=sum, na.rm=T)
names(jh2) <- c('date','state', 'cum.deaths')

jh2.spl <- split(jh2, jh2$state)
jh2.spl <-lapply(jh2.spl, function(x){
  x$deaths<-x$cum.deaths
  for(i in 2:nrow(x)){
    x$deaths[i] <- x$cum.deaths[i] - x$cum.deaths[(i-1)]
  }
  return(x)
})
jh2 <- do.call('rbind.data.frame', jh2.spl)
jh2$state <- state.abb[match(jh2$state, state.name)]

#round date to sunday
jh2$date <-floor_date(jh2$date, 'week')

jh2 <- merge(jh2, hhs_states.cw, by='state')

jh2.agg<- aggregate( jh2$deaths, by=list('state'=jh2$state_region, 'date'=jh2$date), FUN=sum)
names(jh2.agg) <- c('state', 'date', 'deaths')

excess_deaths.df <- cbind.data.frame('date'=dates1,excess_deaths[,,1])
excess_deaths.m <- melt(excess_deaths.df, id.vars =c('date'))
names(excess_deaths.m) <-c('date','state','excess_pi')

#excess_deaths_var
excess_deaths.var.df <- cbind.data.frame('date'=dates1,excess_deaths_var[,,1])
excess_deaths.var.m <- melt(excess_deaths.var.df, id.vars =c('date'))
names(excess_deaths.var.m) <-c('date','state','excess_pi_var')

excess_deaths.upi.df <- cbind.data.frame('date'=dates1,  (obs[,,1]-lpi[,,1] ))
excess_deaths.upi.m <- melt(excess_deaths.upi.df, id.vars =c('date'))
names(excess_deaths.upi.m) <-c('date','state','excess_deaths.upi')

excess_deaths.lpi.df <- cbind.data.frame('date'=dates1,(obs[,,1]-upi[,,1]) )
excess_deaths.lpi.m <- melt(excess_deaths.lpi.df, id.vars =c('date'))
names(excess_deaths.lpi.m) <-c('date','state','excess_deaths.lpi')

total_pi_df <- cbind.data.frame('date'=dates1,(obs[,,1]) )
total_pi_df.m <- melt(total_pi_df, id.vars =c('date'))
names(total_pi_df.m) <-c('date','state','total_pi')

#states.cdc <- unique(excess_deaths.m$state)
jh3 <- merge(jh2.agg, excess_deaths.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.lpi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.upi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, total_pi_df.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.var.m, by=c('state','date') , all=T)

#jh3 <-jh3[jh3$state %in% states.cdc,]


jh3 <- merge( jh3, pi.data.clean2[c('state', "week_start",'percent_complete')], by.x=c('state','date') , by.y=c('state',"week_start"), all=T)

jh3 <- jh3[jh3$date >= as.Date('2020-02-09'),] #2 weeks before 1st reported deaths by CDC
jh3$deaths[is.na(jh3$deaths)] <- 0

jh3 <- merge(jh3, test.ds3.spl, by=c('state','date'), all=T)
jh3$covid.track.death[is.na(jh3$covid.track.death)] <- 0

jh3 <- jh3[ !is.na(jh3$deaths), ]

#Only plot out to 1 week ahead of the US P&I data
jh3 <- jh3[ jh3$date <= max(pi.data$"week_start" +7),]
jh3 <- jh3[jh3$state %in% states.cdc,]
#jh3 <-jh3[!is.na(jh3$excess_pi ),]
count.excess <-jh3[jh3$date >= as.Date('2020-02-09') & jh3$date <=max(pi.data$week_end) ,]

count.excess$excess_pi_adj <-count.excess$excess_pi/count.excess$percent_complete

count.excess$total_pi_adj <-count.excess$total_pi/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi','total_pi','covid.track.death','excess_pi_var')], by=list('state'=count.excess$state) ,FUN=sum)

count.excess2 <- merge(count.excess2, pop2, by='state')

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi + 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi - 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_ucl_INC <- (count.excess2$excess_pi + 1.96*sqrt(count.excess2$excess_pi_var))/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_ci_lcl_INC <- (count.excess2$excess_pi - 1.96*sqrt(count.excess2$excess_pi_var))/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_pi_INC <- count.excess2$excess_pi/count.excess2$census_bureau_pop_2019*1000000

count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 
count.excess2$excess_pi_ci_INC <- paste0(round(count.excess2$exces_pi_INC,1), ' (',  round(count.excess2$exces_ci_lcl_INC,1), ', ', round(count.excess2$exces_ci_ucl_INC,1),')'  ) 

#sort by incidence
count.excess2 <- count.excess2[rev(order(count.excess2$exces_pi_INC)),]

count.excess3.unadjusted <- count.excess2[, c('state','total_pi','excess_pi_ci' ,'covid.track.death','excess_pi_ci_INC')]
names(count.excess3.unadjusted) <-c('State','Total P&I deaths', 'Excess P&I deaths', 'Reported COVID-19 deaths','Excess P&I deaths/1,000,000 people')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")
compare.excess <- merge(count.excess3.unadjusted[,c('State',"Excess P&I deaths")], count.excess3[,c('State',"Excess P&I deaths")], by='State')

htmlTable(count.excess3.unadjusted, caption=paste0('Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from February 9, 2020 through ', last.date.format, ' without influenza adjustment in baseline'), rnames=F)
```


