---
title: 'Estimating the early death toll of COVID-19 in the United States'
author: "Dan Weinberger"
date: "4/3/2020"
output:
  html_document:
    df_print: paged
    html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo=F,
  warning=FALSE, 
  message=FALSE,
  comment = "#>"
)
```

```{r setup}
library(ExcessILI)
library(cdcfluview)
library(reshape2)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(plotly)
library(MMWRweek)
library(readr)
library(rjson)
library(htmlTable)
library(RSocrata)
source('./functions/ts_plot_func.R')
```

## Introduction

The COVID-19 outbreak  first emerged in December 2019 in Wuhan, China and rapidly grew into a large-scale global pandemic (WHO pandemic declaration). On March 11, 2020 the World Health Organization labeled the ongoing outbreak a pandemic. For all countries impacted by the pandemic, tracking the severity and impact of COVID-19 illness is, at the time of writing, a critical need, hampered by testing issues and reporting lags for key epidemiological indicators. 

Many countries, including the US, were caught off-guard by the rapidity with which COVID-19 spread from China. Without adequate capacity to test for COVID-19 infection for much of February and March 2020, available laboratory-confirmed cases captured only an estimated 10-15% of all infections.  Although most countries have adopted the strategy to preferentially test severe cases, estimating the number of severe infections and deaths caused by COVID-19 will be a challenge. Typically, a large proportion of deaths caused by infectious diseases are not attributed to a specific pathogen. With the limited availability of testing for the novel coronavirus and with imperfect sensitivity of the tests, there have undoubtedly been a number of deaths caused by the virus that are not counted in official tallies of the death toll of the epidemic. Further, in the midst of a large outbreak, there is an unavoidable delay in compilation of death certificates and ascertainment of cause of deaths, which contributes to the “fog of war”. And even in situations of ample testing, deaths from viral pathogens, including COVID-19, can occur indirectly via secondary bacterial infections or exacerbation of chronic conditions . In these indirect pathways, the presence of the virus near the time of death is unlikely and hence undetectable by PCR.

To estimate the burden of death due to a novel respiratory pathogen, previous studies have compared the observed incidence of deaths ascribed to pneumonia and influenza (“P&I”) with the baseline incidence of P&I that would be expected at that time of year. These “excess deaths” provide a crude estimate of pathogen-specific burden, and this approach was used in the early months of the 2009 influenza H1N1-pdm pandemic. Following the 2009 pandemic, it was estimated that just 1 in  7 deaths in the US caused by influenza were confirmed by laboratory intesting (Simonsen et al, Plos Med 2013 (says so in the abstract --  link https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1001558, Charu et al IORV 2013). 

In this study, we estimated the increase in P&I deaths across the United States in each week in excess of a seasonal baseline to capture the direct and indirect mortality burden of COVID-19. We compare these estimates of excess deaths to the reported numbers of deaths due to COVID-19 in different states and evaluate the trajectories of these increases in relation to increases in influenza-like illness that could be linked to COVID-19. These analyses provide insights into the true burden of COVID-19 in the early months of the outbreak in the United States.

##METHODS
**Data**
Data on deaths due to pneumonia and influenza by state and week were obtained from the National Center for Health Statistics’ (NCHS) mortality surveillance system. The P&I mortality data provide the provisional number of deaths due to pneumonia or influenza and the total number of deaths in each week. The mortality data are available with a 2 week lag. 

State-level data on influenza-like illness (ILI) in each week were obtained from the CDC’s ILINet system, which aggregates data from a network of outpatient providers. State-level data on influenza activity were obtained from the CDC’s National Respiratory and Enteric Virus Surveillance System (NREVSS). This dataset captures the number of tests performed for influenza and the number that were positive by week and state. The ILI data, NREVSS data, and P&I mortality data were accessed through the CDC’s FluView portal using the cdcfluview package in R. The ILI data provide the percent of visits to participating outpatient providers that were for ILI. ILINet and NREVSS data are available with a 1-week lag. 
Data on reports of laboratory-confirmed deaths due to COVID-19 in each week and state were obtained from several sources, including the covidtracking project, the Hopkins database, and NCHS XX. https://data.cdc.gov/NCHS/Provisional-Death-Counts-for-Coronavirus-Disease-C/hc4f-j6nb  
https://covidtracking.com/ State-specific testing information was obtained was https://covidtracking.com/

**Analysis**
We fit Poisson regression models to the weekly state level data on deaths due to pneumonia and influenza. We adjusted for seasonality using harmonics variables and including year-to-year variations using dummy variables. Influenza activity was adjusted for by adjusting for the percent of tests that were positive for influenza in the previous week. The 1 week lag between the testing data and the mortality data is included to account for lags between when a person is tested for influenza and when they die. The same model was fit to data on influenza-like illness. As a sensitivity analysis, we refit the models, excluding the influenza terms to account for possible biases in influenza testing occurring during the COVID-19 epidemic. To calculate the excess number of deaths, we subtracted the expected number of deaths in each week from the observed number of deaths and divided this by the proportion completeness. This correction was used to adjust for under-reporting in recent weeks.   


## RESULTS


```{r}
## Download the mortality data
#pi.data.alt <- read.socrata("https://data.cdc.gov/resource/pp7x-dyj2.json")
#pi.data.alt <- pi.data.alt[pi.data.alt$geoid=='state',]
pi.data <- pi_mortality(coverage_area='state')
pi.data$percent_complete[pi.data$percent_complete > 1] <- 1
```

```{r}
plot.states <- c('CA', 'FL','GA','ID','IL', 'LA','MA','MI','MO', 'MT', 'NJ', 'NY', 'OR', 'PA', 'WA' )
```

```{r}
#TESTING DATA FORMATTING
url.test<-"https://covidtracking.com/api/v1/states/daily.json" 


json_data <- fromJSON(file=url.test)
test.dates <- as.character(sapply(json_data,'[[','date'))

test.state <- sapply(json_data,'[[','state')
testN <- sapply(json_data, '[[', 'totalTestResults')
cum.death <- sapply(json_data, '[[', 'death')
cum.death <-sapply(cum.death, function(x){ 
  if(is.null(x)){
  x <-0
  }
  return(x)
  }
  )
test.ds <- cbind.data.frame('state'=test.state,test.dates, 'cum.tests'=testN, 'cov.track.cum.death'=cum.death )
test.ds$test.dates <-as.Date(test.ds$test.dates, '%Y%m%d')
test.ds$dow <- weekdays(test.ds$test.dates)

#Look at number of test at end of week..match this to the start date
test.ds <- test.ds[test.ds$dow == 'Saturday' ,]
test.ds$test.dates <- test.ds$test.dates - 6

pop1<-read.csv('./Data/nst-est2019-01.csv')
pop1$state_name <- substring(pop1$state_name,2)
pop1$state <- state.abb[match(pop1$state_name, state.name)]
pop1$census_bureau_pop_2019 <- gsub( ',', '',pop1$census_bureau_pop_2019)
pop1$census_bureau_pop_2019 <-as.numeric(pop1$census_bureau_pop_2019)

test.ds2 <- merge(test.ds, pop1, by='state')
test.ds2$cum.test.per.capita <- test.ds2$cum.tests/test.ds2$census_bureau_pop_2019*1000
test.ds2 <- test.ds2[, c('state','cum.test.per.capita', 'test.dates','cov.track.cum.death')]

test.ds2.spl <- split( test.ds2, test.ds2$state)
test.ds2.spl <- lapply(test.ds2.spl, function(x){
  if(nrow(x)>0){
  x<-x[order(x$test.dates),]
  x$test.week.per.capita <- x$cum.test.per.capita
  x$covid.track.death    <- x$cov.track.cum.death
  
  for(i in 2: nrow(x)){
    x$test.week.per.capita[i] <- x$cum.test.per.capita[i] - x$cum.test.per.capita[i-1]
    
    x$covid.track.death[i] <- x$cov.track.cum.death[i] - x$cov.track.cum.death[i-1]
  }
  }
  return(x)
}
  )
test.ds3.spl <-do.call('rbind.data.frame', test.ds2.spl)
```

```{r}
#download the NREVSS data
nrevvs.state <- cdcfluview::who_nrevss(region = c("state"))
  
  clin <- nrevvs.state[["clinical_labs"]]
  
  data(cdcfluview::hhs_regions)
  
  cw.file <- cdcfluview::hhs_regions
  
  clin2 <- merge(clin, cw.file,
                 by.x = "region",
                 by.y = "state_or_territory")
  
  clin2.subsetvars <- 
    c('region', 'region_number',
      'year', 'week', 'wk_date',
      'total_a','total_b',
      'total_specimens')
  
  clin2 <- clin2[, clin2.subsetvars]
  
  names(clin2)[1:2] <- c("state", "hhs_region")
  
  nrevvs_hhs <- cdcfluview::who_nrevss(region = c("hhs"))
  
  clin.hhs <- nrevvs_hhs[["clinical_labs"]]
  clin.hhs.subsetvars <-
    c('region',
      'wk_date',
      "total_a",'total_b',
      'total_specimens')
  
  clin.hhs <- clin.hhs[, clin.hhs.subsetvars]
  clin.hhs$region <- as.numeric(gsub("Region ", "", clin.hhs$region))
  
  names(clin.hhs) <-
    c("hhs_region",
      "wk_date",
      "hhs_total_a",'hhs_total_b',
      'hhs_total_specimens')
  
  clin3 <- merge(clin2, clin.hhs,
                 by = c("hhs_region", "wk_date"))
  
  clin3$total_a[is.na(clin3$total_a)] <-
    clin3$hhs_total_a[is.na(clin3$total_a)]
  
  clin3$total_b[is.na(clin3$total_b)] <-
    clin3$hhs_total_b[is.na(clin3$total_b)]
  
  clin3$total_specimens[is.na(clin3$total_specimens)] <-
    clin3$hhs_total_specimens[is.na(clin3$total_specimens)]
  
  clin3$state.abb <- state.abb[match(clin3$state, state.name)]
  
  names(clin3) <-
    c("hhs_region",
      "wk_date",
      "state_name",
      "MMWRyear", "MMWRweek",
      "total_a",'total_b',
      'total_specimens',
      'total_a_hhs', "total_b_hhs",
      'total_specimens_hhs',
      "state")
  
  clin3$total_a         <- as.numeric(clin3$total_a)
  clin3$total_b         <- as.numeric(clin3$total_b)
  clin3$total_specimens <- as.numeric(clin3$total_specimens)
  clin3$flu_pct_adj     <- (clin3$total_a + clin3$total_b + 0.5) / 
    (clin3$total_specimens + 0.5)
  clin3$fluN            <- clin3$total_a + clin3$total_b + 0.5
  clin3$flu.var         <- clin3$flu_pct_adj
  
  clin4<-clin3[,c('state','flu_pct_adj', 'wk_date')]
  
  clin4.lag1<-clin4
  clin4.lag1$wk_date <- clin4$wk_date + days(7)
  names(clin4.lag1) <-c('state','flu_pct_adj_lag1','wk_date')
  
  clin4.lag2<-clin4
  clin4.lag2$wk_date <- clin4$wk_date + days(14)
   names(clin4.lag2) <-c('state','flu_pct_adj_lag2','wk_date')
   
clin4.lags <- merge(clin4, clin4.lag1, by=c('state','wk_date'))
clin4.lags <- merge(clin4.lags, clin4.lag2, by=c('state','wk_date'))


```

```{r, include=F}
#Format and fill mssings with 0s

pi.data$state <- state.abb[match(pi.data$region_name, state.name)]
pi.data$state[pi.data$region_name == 'New York City'] <- 'NYC'
spl1<-split(pi.data, pi.data$state)
min.state <- lapply(spl1, function(x){ x$miss.x<-min(x$total_pni)
return(x)
                })
pi.data.clean <- do.call('rbind.data.frame',min.state)
pi.data.clean <- pi.data.clean[!is.na(pi.data.clean$miss.x),]

pi.data.clean2<- merge( pi.data.clean,clin4.lags, by.x=c('week_start', 'state'), by.y=c('wk_date','state'))

pi.data.clean2<- pi.data.clean2[order(pi.data.clean2$state, pi.data.clean2$week_start),]
states.cdc <- unique(pi.data.clean2$state)

```

```{r, eval=F}
#What is most appropriate lag to use for NREVSS data?
pi.data.clean2.spl <- split(pi.data.clean2, pi.data.clean2$state)
cor.lags <- sapply(pi.data.clean2.spl, function(x){
   cor(x[,c("percent_pni","flu_pct_adj",'flu_pct_adj_lag1','flu_pct_adj_lag2')])
 }, simplify='array')
matplot(cor.lags['percent_pni',-1,], type='l')

#Is lag of 0,1,or 2 the best?...shows 1 weeks lag best in 20 states,  2 week lag best in 3 states
table(apply(cor.lags['percent_pni',-1,],2, function(x) which(max(x)==x)))


```

```{r, include=F}
#Run analysis
excess_deaths1.adjusted <-
  excessCases(ds = pi.data.clean2,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "flu_pct_adj_lag1",
              #covs=c("flu_pct_adj", "flu_pct_adj_lag1", "flu_pct_adj_lag2"),
              use.syndromes = c("total_pni"),
              extrapolation.date = "2020-03-01",
              time.res='week')
```


```{r, echo=F}
 #dashboardPlot(excess_deaths1.adjusted)
```


```{r}
### Extract the quantities of interest

#Which syndrome do you want to plot, and over what time range?

syndrome.select <- 'total_pni' #which syndrome do you want to plot?
n.days<-52  #How many days to plot?
ds <- excess_deaths1.adjusted
```
 
```{r}
#Extract the data needed to plot from the results

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred<-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr <-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

excess_deaths <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")
```

## Figure 1: Observed weekly death rate vs seasonal baseline (+/-95% Prediction Interval) 
 
```{r}
n.days <- 52
select.indices <- (length(dates1)-n.days):length(dates1)
dates<-dates1[select.indices]
states <- dimnames(pred)[[2]]
ages <- dimnames(pred)[[3]]
```

The black line shows the observed proportion of deaths that were due to Pneumonia & Influenza (P&I) per week. The red line and shaded area represent the 95% Prediction Interval. The latest data is for the week ending `r  max(dates1)+7`.

```{r, fig.width=6, fig.height=8}

  par(mfrow=c(5,3), mar=c(2,4,2,1))

plot.state.indices <- which(dimnames(pred)[[2]] %in% plot.states )
for(i in plot.state.indices){
  for(j in 1:dim(pred)[3]){
    y.range1<-range(c( pred[select.indices,i,j]/denom[select.indices,i,j],obs[select.indices,i,j]/denom[select.indices,i,j], upi[select.indices,i,j]/denom[select.indices,i,j],0))
      state.name.plot <-  state.name[match(states[i],state.abb)]

  plot(dates,
       pred[select.indices,i,j]/denom[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       xlab='',
       ylab='Proportion',
       #main=paste(states[i])
       )
  
  text(x=as.Date('2019-03-01'),y=0.005, state.name.plot, pos=4)

  points(dates,
         obs[select.indices,i,j]/denom[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates,
            rev(dates)),
          c(lpi[select.indices,i,j]/denom[select.indices,i,j],
            rev(upi[select.indices,i,j]/denom[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

## Figure 2: Observed deaths/expected deaths by state

```{r, echo=F}
rr2<-rr[,,1]
date.mmwrdates <- mmwr_week(dates1)
mmwr.epiyr<- date.mmwrdates$mmwr_year
mmwr.epiyr[date.mmwrdates$mmwr_week<=26] <- mmwr.epiyr[date.mmwrdates$mmwr_week<=26] - 1

mmwr.epiwk <- date.mmwrdates$mmwr_week
mmwr.epiwk[date.mmwrdates$mmwr_week>=27]<-date.mmwrdates$mmwr_week[date.mmwrdates$mmwr_week>=27] - 52
mmwr.epiwk <- mmwr.epiwk +26
check<-cbind.data.frame(date.mmwrdates,mmwr.epiwk, mmwr.epiyr)
```

These plots show the Observed/Expected number of deaths due to pneumonia and influenza in each week for the 2019-20 year (red) compared to previous years (gray). Values close to 1 indicate that the values for that week are close to what would be expected based on the time of year and influenza activity.

```{r, fig.width=6, fig.height=8}

  par(mfrow=c(5,3), mar=c(2,4,2,1))
plot.state.indices <- which(dimnames(pred)[[2]] %in% plot.states )

for(i in plot.state.indices){
    y.range1<-c(0,2)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    matplot(ds2.c$epiwk         ,
       exp(ds2.c[,-1]),
       type='l',
       col=cols1,
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected',
       main=paste(states[i]))
    abline(h=1, col='black')
    }
```



```{r, echo=F, eval=F}
## excess deaths

excess_deaths2 <- excess_deaths[dates1 >= as.Date('2020-02-02'),,1]
excess_deaths.state <- apply(excess_deaths2,2,sum)
cumsum_excess_deaths_state <- apply(excess_deaths2,2,cumsum)
#matplot(cumsum_excess_deaths_state, type='l', bty='l')

```

## Figure 3: Reported number of COVID-19 deaths,compared with the excess deaths due to pneumonia and influenza in each week, by state. 
The blue line shows the number of excess P&I cases, the red solid line shows the reported number of COVID-19 deaths for the same week (as compiled by the Johns Hopkins CSEE), and the dotted red line shows the reported COVID-19 deaths for weeks in which the CDC data were not yet available. Excess P&I estimates from recent weeks are adjusted for incomplete records using the "percent_complete" variable.

```{r}
jh1.url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
jh1<-as.data.frame(read_csv(url(jh1.url)))
jh1 <- jh1[jh1$iso2=='US',]
jh1.m <- melt(jh1[, c(6:7,13:ncol(jh1))], id.vars=c('Admin2',"Province_State"))
jh1.m$variable <-as.Date(jh1.m$variable, '%m/%d/%y')

#Aggregate across the admin2 regions by day
jh2 <- aggregate( jh1.m$value, by=list('date'=jh1.m$variable, 'state'=jh1.m$Province_State), FUN=sum, na.rm=T)
names(jh2) <- c('date','state', 'cum.deaths')

jh2.spl <- split(jh2, jh2$state)
jh2.spl <-lapply(jh2.spl, function(x){
  x$deaths<-x$cum.deaths
  for(i in 2:nrow(x)){
    x$deaths[i] <- x$cum.deaths[i] - x$cum.deaths[(i-1)]
  }
  return(x)
})
jh2 <- do.call('rbind.data.frame', jh2.spl)
jh2$state <- state.abb[match(jh2$state, state.name)]

#round date to sunday
jh2$date <-floor_date(jh2$date, 'week')

jh2.agg<- aggregate( jh2$deaths, by=list('state'=jh2$state, 'date'=jh2$date), FUN=sum)
names(jh2.agg) <- c('state', 'date', 'deaths')

excess_deaths.df <- cbind.data.frame('date'=dates1,excess_deaths[,,1])
excess_deaths.m <- melt(excess_deaths.df, id.vars =c('date'))
names(excess_deaths.m) <-c('date','state','excess_pi')

excess_deaths.upi.df <- cbind.data.frame('date'=dates1,  (obs[,,1]-lpi[,,1] ))
excess_deaths.upi.m <- melt(excess_deaths.upi.df, id.vars =c('date'))
names(excess_deaths.upi.m) <-c('date','state','excess_deaths.upi')

excess_deaths.lpi.df <- cbind.data.frame('date'=dates1,(obs[,,1]-upi[,,1]) )
excess_deaths.lpi.m <- melt(excess_deaths.lpi.df, id.vars =c('date'))
names(excess_deaths.lpi.m) <-c('date','state','excess_deaths.lpi')

total_pi_df <- cbind.data.frame('date'=dates1,(obs[,,1]) )
total_pi_df.m <- melt(total_pi_df, id.vars =c('date'))
names(total_pi_df.m) <-c('date','state','total_pi')

#states.cdc <- unique(excess_deaths.m$state)
jh3 <- merge(jh2.agg, excess_deaths.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.lpi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.upi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, total_pi_df.m, by=c('state','date') , all=T)
jh3 <-jh3[jh3$state %in% states.cdc,]

jh3 <- merge( jh3, pi.data[c('state', "week_start",'percent_complete')], by.x=c('state','date') , by.y=c('state',"week_start"), all=T)

jh3 <- jh3[jh3$date >= as.Date('2020-01-01'),]
jh3$deaths[is.na(jh3$deaths)] <- 0

jh3 <- merge(jh3, test.ds3.spl, by.x=c('state','date'), by.y=c('state','test.dates'), all=T)
jh3$covid.track.death[is.na(jh3$covid.track.death)] <- 0

jh3 <- jh3[ !is.na(jh3$deaths), ]

#Only plot out to 1 week ahead of the US P&I data
jh3 <- jh3[ jh3$date <= max(pi.data$"week_start" +7),]

```

```{r, fig.width=6, fig.height=6}
#par(mfrow=c(5,3), mar=c(2,3,1,2))
#could use 'states.cdc' or plot.states
#ts.plot.func(ds.plot=jh3, death.var='deaths' #,states.plot=plot.states )


```

```{r, fig.width=6, fig.height=6}
## Alternate Fig 3: covidtracking.com data

par(mfrow=c(5,3), mar=c(2,4,1,2))
ts.plot.func(ds.plot=jh3, death.var='covid.track.death' ,states.plot=plot.states )


```

## Table 1

```{r}
count.excess <-jh3[jh3$date >= as.Date('2020-01-01') & jh3$date <=max(pi.data$week_end) &jh3$state %in% states.cdc,]

count.excess$excess_pi_adj <-count.excess$excess_pi/count.excess$percent_complete

count.excess$total_pi_adj <-count.excess$total_pi/count.excess$percent_complete

count.excess$approx.var <- ((count.excess$excess_deaths.upi -count.excess$excess_pi)/2)^2/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi','total_pi','covid.track.death','approx.var')], by=list('state'=count.excess$state) ,FUN=sum)

count.excess2 <- merge(count.excess2, pop1, by='state')

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi + 1.96*sqrt(count.excess2$approx.var))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi - 1.96*sqrt(count.excess2$approx.var))

count.excess2$exces_ci_ucl_INC <- count.excess2$excess_pi + 1.96*sqrt(count.excess2$approx.var)/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_ci_lcl_INC <- count.excess2$excess_pi - 1.96*sqrt(count.excess2$approx.var)/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_pi_INC <- count.excess2$excess_pi/count.excess2$census_bureau_pop_2019*1000000

count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 
count.excess2$excess_pi_ci_INC <- paste0(round(count.excess2$exces_pi_INC,1), '(',  round(count.excess2$exces_ci_lcl_INC,1), ', ', round(count.excess2$exces_ci_ucl_INC,1),')'  ) 

#sort by count
count.excess2 <- count.excess2[rev(order(count.excess2$excess_pi)),]


count.excess3 <- count.excess2[, c('state','total_pi','excess_pi_ci' ,'excess_pi_ci_INC','covid.track.death')]
names(count.excess3) <-c('State','Total P&I deaths', 'Excess P&I deaths', 'Excess P&I deaths/1,000,000 people','Reported COVID019 deaths')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")

htmlTable(count.excess3, caption=paste0('Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from January 5, 2020 through ', last.date.format))
```


```{r}
## And finally,  use the NCHS COVID death data
#(Not yet available with state-level resolution)
#cdc.covid.url<- "https://data.cdc.gov/resource/hc4f-j6nb.json"
```


## Figure 4: Compare Excess P&I mortality vs Excess ILI

Here we compare the observed vs expected number of deaths due to pneumonia and influenza in each week compare to the observed vs expected number of outpatient visits for influenza-like illness (ILI) in each week. we would expect ILI (blue line) to increase earlier than deaths (red line)
```{r, include=F}
ili.data <- ilinet(region = c("state"))
ili.data$state <- state.abb[match(ili.data$region, state.name)]
ili.data       <- ili.data[, c("state", "week_start", "ilitotal", "total_patients")]
ili.data       <- ili.data[!is.na(ili.data$total_patients),]
ili.data.spl   <- split(ili.data, ili.data$state)

min<-sapply(ili.data.spl, function(x)  min(x$total_patients))

state.select<-names(min)[which(min>0) ]
ili.data <- ili.data[ili.data$state %in% state.select,]
## Run the main analysis function, adjusting for flu using NREVSS data
excess_cases1 <-
  excessCases(ds = ili.data,
              datevar       = "week_start", 
              statevar      = "state",
              denom.var     = "total_patients",
              adj.flu       = "auto",
              use.syndromes = c("ilitotal"),
              extrapolation.date = "2020-03-01",
              time.res='week')

dates.ili <-
  excess_cases1[[1]][[1]][[1]]$date
  
rr.ili <-  excessExtract(ds = excess_cases1,
                syndrome = "ilitotal",
                extract.quantity = "resid1")
rr2.ili <- rr.ili[,,1]

date.mmwrdates.ili <- mmwr_week(dates.ili)
mmwr.epiyr.ili<- date.mmwrdates.ili$mmwr_year
mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] <- mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] - 1

mmwr.epiwk.ili <- date.mmwrdates.ili$mmwr_week
mmwr.epiwk.ili[date.mmwrdates.ili$mmwr_week>=27]<-date.mmwrdates.ili$mmwr_week[date.mmwrdates.ili$mmwr_week>=27] - 52
mmwr.epiwk.ili <- mmwr.epiwk.ili +26
```

```{r, fig.width=8, fig.height=7}
common.states <- intersect(colnames(rr2), colnames(rr2.ili))
rr2.comp <- rr2[,common.states]
rr2.ili.comp <- rr2.ili[,common.states] 
  par(mfrow=c(4,4))

for(i in 1:length(common.states)){
    y.range1<-c(0,2)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2.comp[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    plot(ds2.c$epiwk         ,
       exp(ds2.c[,'2019']),
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected',
       main=paste(common.states[i], ' Deaths(red),','ILI(blue)' ))

  es2<-cbind.data.frame('epiwk'=mmwr.epiwk.ili,'epiyr'=mmwr.epiyr.ili, rr=rr2.ili.comp[,i])
    es2.c<-dcast(es2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
     points(es2.c$epiwk         ,
       exp(es2.c[,'2019']), type='l', col='blue')
    abline(h=1, col='black')
}

```
## DISCUSSION
Monitoring trends in non-specific causes of death, like P&I, provides a window into the true burden of COVID-19. Given the lack of adequate testing and geographical variability in testing, this type of monitoring provides key information on the severity of the epidemic in different geographic regions. This also provides some indication of the degree to which viral testing is missing deaths due to COVID-19. Our findings suggest that the degree to which deaths due to COVID-19 are being correctly attributed to the virus varies by state. Some states, such as Washington, attributed most of their excess P&I deaths to COVID. In contrast, other states, such as Florida and XX, might have missed deaths early on and might be undercounting deaths by a substantial degree currently. 
	Non-specific syndromic endpoints, such as deaths due to P&I, outpatient visits for influenza-like illness, and emergency department visits for fever, can provide a crude but informative measure of the progression of the outbreak. These measures themselves can be biased by changes in how diseases or deaths are recorded. However, in the absence of widespread and systematic testing, they provide a useful measure of epidemic progression and could be informative in tracking how interventions like social distancing influence the course of the epidemic.
	The present COVID-19 pandemic is different from the last four influenza A pandemics, in that the mean age at death is about 80 years.  In this sense, the COVID pandemic is  more like seasonal influenza where most deaths are among elderly, but contrast with the age profile of the 2009 influenza pandemic where the mean age was 37yrs..  Because of this it makes sense to look at complex metrics in addition to estimating the number of excess deaths.  For example, years of life lost. See here https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2843747/
 	It is estimated that the epidemic in the US lags that in Europe by 1-2 weeks. Several European countries have experienced high death toll, particularly Italy and Spain. Consolidated excess mortality estimates are not yet available although real-time total mortality data from the EuroMomo project offers a glimpse of the total burden of the epidemic…. We see….
 	
## Supplementary figure: Excess all-cause deaths

Now instead of modeling the proportion of deaths that were due to pneumonia and influenza, we will model all-caused deaths
```{r}
#Run analysis
pi.data.clean2$one <- 1
pi.data.clean2$percent_complete.adj <-pi.data.clean2$percent_complete - 0.5
excess_deaths1.ac <-
  excessCases(ds = pi.data.clean2,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "percent_complete.adj",
              adj.flu       = "flu_pct_adj_lag1",
              #covs=c("flu_pct_adj", "flu_pct_adj_lag1", "flu_pct_adj_lag2"),
              use.syndromes = c("all_deaths"),
              extrapolation.date = "2020-03-01",
              time.res='week')
```

```{r}
ds <- excess_deaths1.ac
syndrome.select <- 'all_deaths'
excess_deaths.ac <- excessExtract(ds = excess_deaths1.ac,
                syndrome = 'all_deaths',
                extract.quantity = "unexplained.cases")
dates1.ac <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred.ac<-  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr.ac <-  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

```

```{r}
# compare excess all-cause deaths with excess P&I deaths

dates.select <- which(dates1 >= as.Date('2020-01-01'))
#plot( excess_deaths[dates.select,,1],excess_deaths.ac[dates.select,,1])

```

The black line shows the observed number of *all* deaths per week, regardless of cause. The red line and shaded area represent the 95% Prediction Interval. The latest data is for the week ending `r  max(dates1)+7`. Note that these are **adjusted** for percent completeness of the data. There are no obvious jumps in all-cause mortality, even in Washington State

```{r, fig.width=8, fig.height=7}

  par(mfrow=c(4,4))

for(i in 1:dim(pred.ac)[2]){
  for(j in 1:dim(pred.ac)[3]){
    y.range1<-range(c( pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j], upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],0))
  plot(dates,
       pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       ylab='Proportion',
       main=paste(states[i],ages[j]))

  points(dates,
         obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates,
            rev(dates)),
          c(lpi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
            rev(upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

## Supp Figure: Extract official data directly from FL 
This shows that the Johns Hopkins data  are off by ~ 1 week in the date assignments compared to Florida official stats. The Florida data records the 'date case counted', so it is possible this date is not date of death but rather the date when the case was recognized.

```{r}
library(pdftools)
library(readr)
library(gsubfn)
fl_text <-pdf_text("./Data/covid-19-data---daily-report-2020-04-06-1657.pdf")
fl1<-read_lines(fl_text)
death.start <- grep('Death    County         Age Gender', fl1 )
death.start.gap <-death.start +3

fl2 <-fl1[death.start.gap[1] : (death.start.gap[1]+300)]
fl.dates <- strapplyc(fl2, "[0-9/]{8,}", simplify = TRUE)
fl.dates <- unlist(fl.dates)
fl.dates <- as.Date(fl.dates, '%m/%d/%y')
fl3 <-cbind.data.frame('date'=fl.dates,'deaths'=rep(1, length(fl.dates)))
fl3$wkdate <- floor_date(fl3$date, 'week')
fl4 <- aggregate(fl3[,'deaths'], by=list('date'=fl3$wkdate), FUN=sum )
names(fl4) <- c('date', 'official_deaths')

fl5 <- merge(jh3, fl4, by='date', all=T)
fl5 <-fl5[fl5$state=='FL',]
fl5 <-fl5[!is.na(fl5$state),]

par(mfrow=c(1,1))
states.cdc <-'FL'
y2.range.test <- range(fl5$test.week.per.capita, na.rm=T)
for(i in 1:length(states.cdc)){
    ds.select <- fl5[fl5$state==states.cdc[i],]
    y.range1<-range(c(ds.select$covid.track.death, ds.select$excess_pi/ds.select$percent_complete), na.rm=T)
    ds.select$death.early <- ds.select$covid.track.death
    ds.select$death.early[is.na(ds.select$excess_pi)] <- NA
    par(new=FALSE)
    plot(ds.select$date         ,
       ds.select$covid.track.death,
       type='l',
       col='red',
       ylim=c(0,max(y.range1)),
       bty='l',
       lty=3,
       ylab='N Deaths',
       xlab='Date',
       main=paste(states.cdc[i] ))
    
       points(ds.select$date         ,
       ds.select$death.early, type='l', col='red',
       lty=1, lwd=2)
       
      points(ds.select$date         ,
       ds.select$deaths, type='l', col='purple',#JH data
       lty=1, lwd=2)
       
      points(ds.select$date         ,
       ds.select$official_deaths, type='l', col='orange',
       lty=1, lwd=2)

       points(ds.select$date         ,
       ds.select$excess_pi/ds.select$percent_complete, type='l', col='blue', lwd=2)
    abline(h=1, col='black')
    
    par(new=TRUE)
   plot(ds.select$date, ds.select$test.week.per.capita, ylim=y2.range.test, type='l', lty=2, lwd=0.5, col='gray', yaxt='n',xaxt='n', ylab='', xlab='', bty='l')
   axis(side=4,at=c(0,1,2,3,4,5) , labels=F)
 mtext("Tests per 1000", side=4, line=0.2, col='gray', cex=0.75)
    
}

```

## Supp Figure; Fig 1 Observed vs Expected P&I deaths for *all* states
```{r, fig.width=8, fig.height=7}

  par(mfrow=c(4,4))

for(i in 1:dim(pred)[2]){
  for(j in 1:dim(pred)[3]){
    y.range1<-range(c( pred[select.indices,i,j]/denom[select.indices,i,j],obs[select.indices,i,j]/denom[select.indices,i,j], upi[select.indices,i,j]/denom[select.indices,i,j],0))
  plot(dates,
       pred[select.indices,i,j]/denom[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       ylab='Proportion',
       main=paste(states[i],ages[j]))

  points(dates,
         obs[select.indices,i,j]/denom[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates,
            rev(dates)),
          c(lpi[select.indices,i,j]/denom[select.indices,i,j],
            rev(upi[select.indices,i,j]/denom[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```


## Supp figure: Fig 2 for all states.
```{r, fig.width=8, fig.height=7}

  par(mfrow=c(4,4), mar=c(2,4,2,1))
plot.state.indices <- which(dimnames(pred)[[2]] %in% states.cdc )

for(i in plot.state.indices){
    y.range1<-c(0,2)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    matplot(ds2.c$epiwk         ,
       exp(ds2.c[,-1]),
       type='l',
       col=cols1,
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected',
       main=paste(states[i]))
    abline(h=1, col='black')
    }
```

