---
title: "Estimating the early death toll of COVID-19 in the United States"
author: "Dan Weinberger"
date: "4/17/2020"
output:
  html_document:
    df_print: paged
    html_document: null
    toc: yes
    toc_depth: 2
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document: 
    keep_tex:  true
params:
  agg.level: 'state_region'
  n.days.filter: 20
---

```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo=F,
  warning=FALSE, 
  message=FALSE,
  comment = "#>",
  dev=c('png','pdf'),
  fig.path='./figures/'
)
```


```{r, eval=F}
if (!require("devtools")) {
  install.packages("devtools")
}
devtools::install_github("weinbergerlab/ExcessILI")
```

```{r setup}
library(ExcessILI)
library(cdcfluview)
library(reshape2)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(plotly)
library(MMWRweek)
library(readr)
library(rjson)
library(htmlTable)
library(RSocrata)
library(pdftools)
library(readr)
library(gsubfn)
library(dplyr)
library(RCurl)
#library(jsonlite)

source('./functions/ts_plot_func.R')
```

```{r archivfunc}
# Using ExcessILI's data archiving functions, returns the most recent copy of
# output obtained by running a function or formula \code{f}, unless this 
# copy doesn't exist or is older (by modification time) than \code{maxage}.
# In that case, \code{f} is run and the output is archived into the folder
# Data/'storeName' as an RDS file, using the function ExcessILI::storeRDS.
#
# @param storeName A string. The name of the folder to store output in
# @param f A function or formula taking no arguments. Formulas are coerced to
#   functions.
# @param maxage How old can any existing archived file be before \code{f} is 
#   called again?
runIfExpired <- function(storeName, f, maxage=hours(24)) {
  basepath <- "Data/"
  mostRecent <- mostRecentTimestamp(storeName, basepath=basepath)
  f <- rlang::as_function(f)
  
  runAndArchive <- function() {
    data <- f()
    storeRDS(data, storeName, basepath)
    data
  }
    
  if (is.na(mostRecent)) 
    return(runAndArchive())

  if (mostRecent %--% now() < maxage)
    return(retrieveRDS(storeName, basepath))

  runAndArchive()
}
```

```{r states_to_plot}
plot.states <- c('CA','TX','FL', 'NY','PA','IL','GA','MI','LA', 'MA' )

exclude.states <- c('CT', 'NC') #no up-to-date number
```

```{r start_date_select}
#Date when start counting excess deaths: Start first week of Feb to Match CDC's COVIDview tables
count.start.date <- as.Date('2020-01-26')
```


```{r import_state_pi_data}
#Import the P&I data
## Download the mortality data
#pi.data.alt <- read.socrata("https://data.cdc.gov/resource/pp7x-dyj2.json") #only through end of 2019
#pi.data.alt <- pi.data.alt[pi.data.alt$geoid=='state',]

#ARCHIVE
pi.data <- runIfExpired('pi_mortality_state', ~pi_mortality(coverage_area='state'))

pi.data.last <- pi.data[pi.data$week_end==max(pi.data$week_end) & pi.data$geo_description=='State',]

last.date.avail.pi <- 
  max(pi.data$week_end)
last.date.obtained <- 
  last.date.avail.pi +6 #release date fr latest data (if            weekly)

 pi.data <- pi.data[pi.data$week_end <=                     (last.date.obtained-params$n.days.filter),]
 
  pi.data <- pi.data[!(is.na(pi.data$week_end)),]

#pi.data.miss.states <- unique(pi.data.last$region_name[is.na(pi.data.last$total_pni)])
 
#pi.data.miss.states <- unique(pi.data.last$region_name)

pi.data.not.miss.states <- unique(pi.data.last$region_name)

#pi.data.not.miss.states <- unique(pi.data.last$region_name[!is.na(pi.data.last$total_pni)])

pi.data.not.miss.states <- state.abb[match(pi.data.not.miss.states, state.name)]

pi.data.not.miss.states <-
  pi.data.not.miss.states[!is.na(pi.data.not.miss.states)]


```

```{r import_state_adjustments}
#State-level adjustments for under-reporting
adj <- read.csv('./outputs/report.proportion.state.csv')
adj.m <-melt(adj, id.vars='X' )
adj.m$state <- state.abb[match(adj.m$variable, state.name)]
adj.m$state[adj.m$variable=='New York City'] <- 'NYC'

last.pi.date <- max(pi.data$week_end)
adj.m$date <- NA
adj.m$date[adj.m$X==1]  <- as.Date(last.pi.date) 
adj.m$date[adj.m$X==2]  <- as.Date(last.pi.date) - 7
adj.m$date[adj.m$X==3]  <- as.Date(last.pi.date) - 14
adj.m$date[adj.m$X==4]  <- as.Date(last.pi.date) - 21
adj.m$date <- as.Date(adj.m$date, origin='1970-01-01')
adj.m <- adj.m[, c('date', 'state','value')]
names(adj.m) <- c('date', 'state', 'reporting.adjustment')

```

```{r state_region_cw}
#cross walk file to map  states to hhs regions
# 
hhs_states <- cdcfluview::hhs_regions
hhs_states$state <- state.abb[match(hhs_states$state_or_territory, state.name)]
hhs_states$state_region <- hhs_states$region
hhs_states$state_region[hhs_states$state %in% plot.states] <- hhs_states$state[hhs_states$state %in% plot.states]
hhs_states <- hhs_states[,c('state','state_region')]
hhs_states$state_region <- gsub(' ','', hhs_states$state_region)
hhs_states.cw <- hhs_states[!is.na(hhs_states$state),]

#Get rid of states that have no data in recent time points
hhs_states.cw <- 
  hhs_states.cw[ hhs_states.cw$state %in% pi.data.not.miss.states,]

#Exclude CT
hhs_states.cw <- 
  hhs_states.cw[ !(hhs_states.cw$state %in% exclude.states),]


if(params$agg.level== 'state_region'){
hhs_states.cw.spl <- split(hhs_states.cw, 
                            hhs_states.cw$state_region)
}else{
  hhs_states.cw.spl <- split(hhs_states.cw, 
                            hhs_states.cw$state)
}

hhs_states.cw.spl <-lapply(hhs_states.cw.spl, function(x) {
  x$state_region.lab <- paste(x$state, collapse=',')
  return(x)
  }
)

hhs_states.cw <-do.call('rbind.data.frame',hhs_states.cw.spl)
hhs_states.cw$state_region[hhs_states.cw$state_region != hhs_states.cw$state] <-
  hhs_states.cw$state_region.lab[hhs_states.cw$state_region != hhs_states.cw$state]


```

```{r import_covid_tracking}
#TESTING DATA FORMATTING
url.test<-"https://covidtracking.com/api/v1/states/daily.json" 

#ARCHIVE
json_data <- runIfExpired('covidtracking_states_daily', ~fromJSON(file=url.test))
test.dates <- as.character(sapply(json_data,'[[','date'))

test.state <- sapply(json_data,'[[','state')
testN <- sapply(json_data, '[[', 'totalTestResultsIncrease')
testN <-sapply(testN, function(x){ 
  if(is.null(x)){
  x <-0
  }
  return(x)
  }
  )
deathN <- sapply(json_data, '[[', 'deathIncrease')
deathN <-sapply(deathN, function(x){ 
  if(is.null(x)){
  x <-0
  }
  return(x)
  }
  )
test.ds <- cbind.data.frame('state'=test.state,test.dates, 'testN.day'=testN, 'deathN.day'=deathN )
test.ds$test.date.wk <-floor_date(as.Date(test.ds$test.dates, '%Y%m%d'),'week')

test.ds <- merge(test.ds, hhs_states.cw, by='state')


test.ds.agg <- aggregate( test.ds[,c('testN.day','deathN.day')] , by=list('state'=test.ds$state_region,'date'=test.ds$test.date.wk), FUN=sum)
names(test.ds.agg) <- c('state','date','testN','deathN')

pop1<-read.csv('./Data/nst-est2019-01.csv')
pop1$state_name <- substring(pop1$state_name,2)
pop1$state <- state.abb[match(pop1$state_name, state.name)]
pop1$census_bureau_pop_2019 <- gsub( ',', '',pop1$census_bureau_pop_2019)
pop1$census_bureau_pop_2019 <-as.numeric(pop1$census_bureau_pop_2019)

pop1.reg <- merge(pop1, hhs_states.cw, by='state')
pop2 <- aggregate(pop1.reg[,"census_bureau_pop_2019"], by=list('state'=pop1.reg$state_region), FUN=sum)
names(pop2) <- c('state','census_bureau_pop_2019')

test.ds2 <- merge(test.ds.agg, pop2, by='state')
test.ds2$test.week.per.capita <- test.ds2$testN/test.ds2$census_bureau_pop_2019*1000
names(test.ds2) <- c('state','date','testN','covid.track.death', 'pop2019',"test.week.per.capita")

test.ds3.spl <- test.ds2
```

```{r import_cdc_covid_view}
#official CDC data
url.cdc.covid<-"https://data.cdc.gov/resource/hc4f-j6nb.json" 

#ARCHIVE
cdc.data <- runIfExpired('cdc_covid_data', ~fromJSON(file=url.cdc.covid, simplify=F))
cdc.pneum.and.covid <- sapply(cdc.data,'[[', "pneumonia_and_covid_deaths"  )
cdc.total.death <- sapply(cdc.data, '[[', 'total_deaths')
cdc.covid.death <- sapply(cdc.data,'[[', "covid_deaths"  )
cdc.end.wk <- sapply(cdc.data,'[[', "end_week"  )
cdc.pneu.death <- sapply(cdc.data,'[[', "pneumonia_deaths"   )

cdc.grp<-sapply(cdc.data,'[[', "group"  )
cdc.ind<-sapply(cdc.data,'[[', "indicator"  )

cdc.covid.death <-lapply(cdc.covid.death, function(x){ if(is.null(x)){ x<-NA}
return(x)
})
cdc.covid.death <-unlist(cdc.covid.death)

cdc.total.death <-lapply(cdc.total.death, function(x){ if(is.null(x)){ x<-NA}
return(x)
})
cdc.total.death <-unlist(cdc.total.death)

cdc.pneum.and.covid <-lapply(cdc.pneum.and.covid, function(x){ if(is.null(x)){ x<-NA}
return(x)
})
cdc.pneum.and.covid <-unlist(cdc.pneum.and.covid)

cdc.summary <- cbind.data.frame('grp'=cdc.grp,'ind'=cdc.ind,'NCHS Reported COVID-19 Deaths'=cdc.covid.death, 'NCHS Reported COVID-19 Deaths, with pneumonia code'=cdc.pneum.and.covid, 'total deaths'=cdc.total.death, 'week_end'=cdc.end.wk,'pneumonia'=cdc.pneu.death)

cdc.summary.wk <- 
  cdc.summary[cdc.summary$grp=='By week',]

cdc.summary.wk$wk_end <- 
  as.Date(substr(cdc.summary.wk$week_end,1,10), "%Y-%m-%d")

cdc.summary.wk <-
  cdc.summary.wk[!is.na(cdc.summary.wk$wk_end),]

cdc.summary.wk <- cdc.summary.wk[, c('wk_end', 'NCHS Reported COVID-19 Deaths','NCHS Reported COVID-19 Deaths, with pneumonia code','total deaths','pneumonia')]


cdc.summary.state <- cdc.summary[cdc.summary$grp=='By state',]

cdc.summary.state$`NCHS Reported COVID-19 Deaths, with pneumonia code` <- as.numeric(as.character(cdc.summary.state$`NCHS Reported COVID-19 Deaths, with pneumonia code`))

cdc.summary.state$`NCHS Reported COVID-19 Deaths` <- as.numeric(as.character(cdc.summary.state$`NCHS Reported COVID-19 Deaths` ))

cdc.summary.state$covid.pct.pneum <- cdc.summary.state$`NCHS Reported COVID-19 Deaths, with pneumonia code`/cdc.summary.state$`NCHS Reported COVID-19 Deaths`
```


```{r import_national_pic}
## there are 3 sources of national data, each has pluses and minus: 1) NCHS provisional data, which is updated daily and has all deaths and pneumonia deaths and covid deaths but not combined PIC category. this is available since last week of Jan 2) Provisional data from another website on NCHS that is updated weekly and has combined pneumonia/influenza/COVID (PIC) and is avalable since week 40 of 2019 3) The weekly NCHS fluview mortality summary, provides P&I and all cause, but no COVID

nchs_covidview <- cdc.summary.wk

#starting 2019 wk 40, CDC provide combined PIC national
#https://www.cdc.gov/coronavirus/2019-ncov/covid-data/covidview/04172020/nchs-mortality-report.html

#CAUTION: THE PIC data seem to be less complete than the fluview data (even looking at total deaths)
 download.file('https://www.cdc.gov/coronavirus/2019-ncov/covid-data/covidview/04172020/csv/nchs-mortality-report.csv','./Data/pic_provisional/pic.provisional.txt')
 

pic1 <- read.csv('./Data/pic_provisional/pic.provisional.txt', skip=4)
pic1<- pic1[-1,]
pic1 <- apply(pic1,2, function(x) 
  gsub(',','',x))

pic1 <- apply(pic1,2, function(x) 
  as.numeric(as.character(x))) 

pic1 <- as.data.frame(pic1)

pic1 <- pic1[,c("Year", "Week","Total.Deaths","Pneumonia..Influenza.or.COVID.19.Deaths")]
names(pic1) <- c('year', 'week', 'total_deaths_pic_denom','pic_deaths')

nchs_covidview <- 
  nchs_covidview[cdc.summary.wk$wk_end < 
                   max(cdc.summary.wk$wk_end),]

nchs_covidview$`total deaths` <-
  as.numeric(as.character(nchs_covidview$`total deaths`))

nchs_covidview$`pneumonia` <-
  as.numeric(as.character(nchs_covidview$`pneumonia`))

nchs_covidview$`NCHS Reported COVID-19 Deaths` <-
  as.numeric(as.character(nchs_covidview$`NCHS Reported COVID-19 Deaths`))

plot(nchs_covidview$wk_end, nchs_covidview$`total deaths`)

#Also import the latest national fluview data
#natl.all.data <- runIfExpired('PI_mortality_national', #~pi_mortality(coverage_area='national'))
#NOTE can't use national data as baseline bc missing WV, NC, CT
pi.data.exclude <- pi.data
pi.data.exclude$state <-   
  state.abb[match(pi.data.exclude$region_name, state.name)]
pi.data.exclude <- 
  pi.data.exclude[!(pi.data.exclude$state %in% exclude.states),]

natl.all.data <- 
  aggregate(pi.data.exclude[,c("total_pni", 'all_deaths')],
            by=list('week_end'=pi.data.exclude$week_end,
                'year_week_num'=pi.data.exclude$weeknumber ), FUN=sum)


natl.all.data$year <- 
  year(natl.all.data$week_end)

max.wk <- 
  unique(natl.all.data$year_week_num[natl.all.data$week_end==max(pi.data$week_end)])

wk.range <- c(10,max.wk)

#Replace 2020 data with the NCHS data, which are more up to date
#natl.all.data <- 
#  natl.all.data[natl.all.data$year <=2019,]

natl.all.data <- 
  natl.all.data[,c('year','week_end', 'year_week_num',"all_deaths","total_pni" )]

mmwr.wk1 <- MMWRweek(nchs_covidview$wk_end)
nchs_covidview$year_week_num <- mmwr.wk1$MMWRweek
nchs_covidview <- nchs_covidview[,c('wk_end','NCHS Reported COVID-19 Deaths', 'total deaths', 'pneumonia', 'year_week_num')]

names(nchs_covidview) <- 
  c('week_end','covid19.nchs', 'all_deaths_covidview',
    'number_pneumonia','year_week_num')
nchs_covidview <- 
  nchs_covidview[,c('week_end', 'covid19.nchs','all_deaths_covidview')]
#nchs_covidview$year <- 2020

natl.all.data$year_week_num <-
  as.numeric(as.character(natl.all.data$year_week_num))

#natl.all.data <- bind_rows(natl.all.data,nchs_covidview)

natl.all.data <-
  merge(natl.all.data,
        nchs_covidview[,c('week_end','covid19.nchs','all_deaths_covidview')],
        by='week_end', all=T)

#Use 'all-death' column from covid view since this is updated daily
natl.all.data$all_deaths[!is.na(natl.all.data$all_deaths_covidview)] <- 
  natl.all.data$all_deaths_covidview[!is.na(natl.all.data$all_deaths_covidview)]

natl.all.data <-
  merge(natl.all.data, pic1, by.x=c('year','year_week_num'),by.y=c('year','week') ,all=T)

natl.all.data$pic_deaths[is.na(natl.all.data$pic_deaths)] <-
  natl.all.data$total_pni[is.na(natl.all.data$pic_deaths)]

natl.all.data$total_deaths_pic_denom[is.na(natl.all.data$total_deaths_pic_denom)] <-
  natl.all.data$all_deaths[is.na(natl.all.data$total_deaths_pic_denom)]


natl.all.data.sub <-
  natl.all.data[natl.all.data$year_week_num >= wk.range[1] &
                  natl.all.data$year_week_num <=
                  wk.range[2],]

agg.historic <- aggregate(natl.all.data.sub[,c("all_deaths","total_pni", 'covid19.nchs')], by=list('year'=natl.all.data.sub$year), FUN=sum)

natl.all.data$state <- 'NATL'

natl.all.data <-
  natl.all.data[order(natl.all.data$week_end),]
plot(natl.all.data$week_end, natl.all.data$all_deaths, type='l')


nrevvs.natl <- runIfExpired('nrevss_national', ~cdcfluview::who_nrevss(region = c("national")))
nrevvs.natl <- nrevvs.natl$clinical_labs
nrevvs.natl <- nrevvs.natl[,c('wk_date','percent_positive') ]
nrevvs.natl$week_end <- nrevvs.natl$wk_date+6

natl.all.data2 <- merge( natl.all.data, nrevvs.natl, by.x='week_end', by.y='week_end')
```


```{r}
#download the NREVSS data

#ARCHIVE
nrevvs.state <- runIfExpired('nrevss_state', ~cdcfluview::who_nrevss(region = c("state")))
  
  clin <- nrevvs.state[["clinical_labs"]]
  clin$state <- state.abb[match(clin$region, state.name)]
  
  data(cdcfluview::hhs_regions)
  
  cw.file <- cdcfluview::hhs_regions
  
  clin2 <- merge(clin, hhs_states.cw,
                 by = "state")

  clin2 <- merge(clin2, cw.file,
                 by.x = "region",
                 by.y = "state_or_territory")
  
  clin2.subsetvars <- 
    c('region', 'region_number',
      'year', 'week', 'wk_date',
      'total_a','total_b',
      'total_specimens','state_region')
  
  clin2 <- clin2[, clin2.subsetvars]
  
  names(clin2)[1:2] <- c("state", "hhs_region")
  
  clin2$total_specimens <- as.numeric(clin2$total_specimens)
  clin2$total_a <- as.numeric(clin2$total_a)
  clin2$total_b <- as.numeric(clin2$total_b)
  
  clin2 <- aggregate(clin2[, c('total_specimens','total_a','total_b')], by=list('state'=clin2$state_region, 'wk_date'=clin2$wk_date,'hhs_region'=clin2$hhs_region), FUN=sum, na.rm=T)
  
  ##Florida doesn't have ILI data, so use regions ILI dat
  #
  #ARCHIVE
  nrevvs_hhs <- runIfExpired('nrevss_hhs', ~cdcfluview::who_nrevss(region = c("hhs")))
  
  clin.hhs <- nrevvs_hhs[["clinical_labs"]]
  clin.hhs.subsetvars <-
    c('region',
      'wk_date',
      "total_a",'total_b',
      'total_specimens')
  
  clin.hhs <- clin.hhs[, clin.hhs.subsetvars]
  clin.hhs$region <- as.numeric(gsub("Region ", "", clin.hhs$region))
  
  names(clin.hhs) <-
    c("hhs_region",
      "wk_date",
      "hhs_total_a",'hhs_total_b',
      'hhs_total_specimens')
  
  clin3 <- merge(clin2, clin.hhs,
                 by = c("hhs_region", "wk_date"))

  #If no specimens for state, use HHS region estimates
  clin3$total_specimens[clin3$total_specimens==0] <-
    clin3$hhs_total_specimens[clin3$total_specimens==0]
  
  clin3$total_a[clin3$total_specimens==0] <-
    clin3$hhs_total_a[clin3$total_specimens==0]
  
  clin3$total_b[clin3$total_specimens==0] <-
    clin3$hhs_total_b[clin3$total_specimens==0]
  
  clin3$total_specimens <- as.numeric(clin3$total_specimens)
  clin3$total_a <- as.numeric(clin3$total_a)
  clin3$total_b <- as.numeric(clin3$total_b)  
  
  clin3$flu_pct_adj <- (clin3$total_a + clin3$total_b)/clin3$total_specimens
  
   clin4<-clin3[,c('state','flu_pct_adj', 'wk_date')]

  clin4.lag1<-clin4
  clin4.lag1$wk_date <- clin4$wk_date + days(7)
  names(clin4.lag1) <-c('state','flu_pct_adj_lag1','wk_date')
  
  clin4.lag2<-clin4
  clin4.lag2$wk_date <- clin4$wk_date + days(14)
   names(clin4.lag2) <-c('state','flu_pct_adj_lag2','wk_date')
   
clin4.lags <- merge(clin4, clin4.lag1, by=c('state','wk_date'))
clin4.lags <- merge(clin4.lags, clin4.lag2, by=c('state','wk_date'))


```

```{r, include=F}
#Format and fill mssings with 0s

pi.data$percent_complete[pi.data$percent_complete > 1] <- 1

pi.data$state <- state.abb[match(pi.data$region_name, state.name)]

pi.data$state[pi.data$region_name == 'New York City'] <- 'NYC'

pi.data <- pi.data[!(pi.data$state %in% exclude.states),]

pi.data <- merge(pi.data,hhs_states.cw, by='state', all=T)
pi.data$state_region[pi.data$state=='NYC'] <- 'NYC'
pi.data$pct.comp.wgt <- pi.data$percent_complete*pi.data$all_deaths

pi.data.ny.separate <- pi.data

#Combine NYC and NY state
pi.data$state_region[pi.data$state=='NYC'] <- 'NY'


pi.data.agg <- aggregate(pi.data[,c('total_pni','all_deaths','pct.comp.wgt')], by=list('state'=pi.data$state_region, 'week_start'=pi.data$week_start, 'week_end'=pi.data$week_end), FUN=sum, na.rm=T)
pi.data.agg$percent_complete <- pi.data.agg$pct.comp.wgt/pi.data.agg$all_deaths

spl1<-split(pi.data.agg, pi.data.agg$state)
min.state <- lapply(spl1, function(x){ x$miss.x<-min(x$total_pni, na.rm=T)
return(x)
                })
pi.data.clean <- do.call('rbind.data.frame',min.state)
pi.data.clean <- pi.data.clean[!is.na(pi.data.clean$miss.x),]

pi.data.clean2<- merge( pi.data.clean,clin4.lags, by.x=c('week_start', 'state'), by.y=c('wk_date','state'))

states.cdc <- unique(pi.data.clean2$state)

date.print <- max(pi.data.clean2$week_end)

```



### Mortality data on deaths due to pneumonia & influenza through the week ending `r  date.print`

Dan Weinberger,^1^  Ted Cohen,^1^  Forrest W. Crawford,^2^  Farzad Mostashari,^3^  Don Olson,^4^  Virginia E Pitzer,^1^  Nicholas G Reich,^5^  Marcus Russi,^1^ Lone Simonsen,^6^ Annie Watkins,^1^ Cecile Viboud^7^ 

^1^Department of Epidemiology of Microbial Diseases and the Public Health Modeling Unit, Yale School of Public Health, New Haven, CT
^2^Department of Biostatistics and the Public Health Modeling Unit, Yale School of Public Health, New Haven, CT; Yale Departments of Ecology and Evolutionary Biology, Statistics & Data Science, Yale School of Management
^3^Aledade, Inc
^4^Department of Health and Mental Hygiene, New York City, NY
^5^Department of Biostatistics and Epidemiology, School of Public Health and Health Sciences, University of Massachusetts, Amherst, MA
^6^Department of Science and Environment, Roskilde University, Denmark
^7^Division of International Epidemiology and Population Studies, Fogarty International Center, National Institutes of Health, Bethesda, MD

## IMPORTANT NOTE
Starting April 17, fluview started reporting death data with a 1 week lag, rather than a 2 week lag. Data from the recent weeks are highly incomplete and should be interpreted with caution

## Abstract

*Background*
Tracking the severity and public health impact of emerging diseases, such as the novel coronavirus diseases COVID-19, is a critical need. These efforts are often hampered by testing issues and reporting lags for key epidemiological data. Evaluating unexplained increases in deaths attributed to non-specific causes, such as pneumonia, can provide a more complete picture of the burden caused by COVID-19. 

*Methods*
We evaluated increases in the occurrence of deaths due to P&I above a seasonal baseline across the United States and compared the rate of excess deaths to reported rates of death due to COVID-19, as well as state-level coronavirus testing rates and excess rates of influenza-like illness. 

*Results*
There were notable increases in the rate of death due to pneumonia and influenza. In a number of states, these increases pre-dated the increase in testing rates and went uncounted. There was substantial variability between states in the discrepancy between reported rates of death due to COVID-19 and the estimated burden of excess deaths due to P&I. In some states, the burden of reported deaths was 10-fold lower than the burden of excess deaths due to P&I. 

*Conclusions*
Given the lack of consistent virological testing, tracking spikes in deaths due to non-specific causes provides the most complete estimate of the burden of COVID-19. This approach provides a framework for tracking the progression of the epidemic and its public health impact. 

## National analysis of P&I
```{r natl_pi_analysis, include=F}
#Run analysis
natl.all.data2$one <- 1
natl.all.data3 <-
  natl.all.data2[!is.na(natl.all.data2$pic_deaths),]
excess_ac_natl <-
  excessCases(ds = natl.all.data3,
              datevar       = "week_end",
              statevar      = "state",
              adj.flu       = "percent_positive",
              denom.var     = 'total_deaths_pic_denom',
              use.syndromes = c("pic_deaths"),
              extrapolation.date = '2020-03-01',
              model.type='negbin',
              time.res='week')
ds <- excess_ac_natl

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "unexplained.cases")

pred <- 
   excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "pred")

upi <- 
   excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "upi")

lpi <- 
   excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "lpi")

obs <- 
   excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "y")

sum.pred.iter <-    
  excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "sum.pred.iter")

sum.obs <-
    excessExtract(ds = ds,
                syndrome = 'pic_deaths',
                extract.quantity = "sum.obs")

sum.cases.excess <- 
  sum(sum.obs) - sum.pred.iter

sum.excess.deaths.range.pi <-
  quantile(sum.cases.excess[,1,1], probs=c(0.025,0.5,0.975))

plot(natl.all.data3$week_end, natl.all.data3$pic_deaths, type='l')
points(natl.all.data3$week_end,pred, type='l', col='gray')

res.pi1 <-cbind.data.frame('week_end'=natl.all.data3$week_end,
                          'pred'=pred[,1,1], 'upi'=upi[,1,1], 'lpi'=lpi[,1,1],'obs'=obs[,1,1],
                      'unexplained.cases'=unexplained.cases[,1,1])

mmwr.rs <- mmwr_week(res.pi1$week_end)
res.pi1<- cbind.data.frame(mmwr.rs, res.pi1)
#plot(res.pi$week_end, res.pi$unexplained.cases)



res.pi <- res.pi1[res.pi1$week_end<=as.Date('2020-04-04'),]

write.csv(res.pi, './outputs/p_i_natl_obs_expected.csv')

sum(res.pi$unexplained.cases[res.pi$mmwr_week>=10 &
                               res.pi$mmwr_week<=14 & res.pi$week_end>=as.Date('2020-01-01')])
```

Excess deaths due to pneumonia/influenza/covid nationally in the specified time period
```{r}
print(sum.excess.deaths.range.pi)
```

```{r natl_pi_plot}
yrange.pneu <- range(c(res.pi$pred, res.pi$obs, res.pi$upi,0))

plot(res.pi$week_end, res.pi$pred, type='l', ylim=yrange.pneu, col='gray', )

polygon(c(res.pi$week_end, rev(res.pi$week_end)), c(res.pi$lpi, rev(res.pi$upi)),col = rgb(0, 0, 0, alpha = 0.1), border = NA )

points(res.pi$week_end, res.pi$obs, type='l', col='red', bty='l')

```

## National analysis of all-cause 

```{r, include=F}
#Run analysis
natl.all.data2$one <- 365000000/100000
natl.all.data3 <-
    natl.all.data2[!is.na(natl.all.data2$total_pni),]
excess_ac_natl <-
  excessCases(ds = natl.all.data3,
              datevar       = "week_end",
              statevar      = "state",
              adj.flu       = "percent_positive",
              denom.var     = 'one',
              use.syndromes = c("all_deaths"),
              extrapolation.date = '2020-03-01',
              model.type='negbin',
              time.res='week')
ds <- excess_ac_natl

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "unexplained.cases")

pred <- 
   excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "pred")

upi <- 
   excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "upi")

lpi <- 
   excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "lpi")

obs <- 
   excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "y")

sum.obs.ac <-
    excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "sum.obs")

sum.pred.iter.ac <-    
  excessExtract(ds = ds,
                syndrome = 'all_deaths',
                extract.quantity = "sum.pred.iter")

sum.ac.excess <- 
  sum(sum.obs.ac) - sum.pred.iter.ac

sum.excess.deaths.range.ac <-
  quantile(sum.ac.excess[,1,1], probs=c(0.025,0.5,0.975))

plot(natl.all.data2$week_end, natl.all.data2$all_deaths, type='l')
points(natl.all.data2$week_end,pred, type='l', col='gray')

res.ac1 <-cbind.data.frame('week_end'=natl.all.data2$week_end,
                          'pred'=pred[,1,1], 'upi'=upi[,1,1], 'lpi'=lpi[,1,1],'obs'=obs[,1,1],
                      'unexplained.cases'=unexplained.cases[,1,1])

mmwr.rs <- mmwr_week(res.ac1$week_end)
res.ac1<- cbind.data.frame(mmwr.rs, res.ac1)
#plot(res.ac$week_end, res.ac$unexplained.cases)

res.ac <- res.ac1[res.ac1$week_end<=as.Date('2020-04-04'),]

yrange.pneu <- range(c(res.ac$pred, res.ac$obs, res.ac$upi,0))

sum(res.ac$unexplained.cases[res.ac$mmwr_week>=10 & res.ac$mmwr_week<=14 & res.ac$week_end>=as.Date('2020-01-01')])

```

Total excess all cause deaths
```{r}
print(sum.excess.deaths.range.ac)
```


```{r plot_ac_natl}
plot(res.ac$week_end, res.ac$pred, type='l', ylim=yrange.pneu, col='gray')

polygon(c(res.ac$week_end, rev(res.ac$week_end)), c(res.ac$lpi, rev(res.ac$upi)),col = rgb(0, 0, 0, alpha = 0.1), border = NA )

points(res.ac$week_end, res.ac$obs, type='l', col='red', bty='l')
```

```{r summary_table_natl}
res.pi2 <- res.pi1
names(res.pi2) <-
  c('year','week','day','week_end_date','baseline_pi','baseline_pi_upper', 'baseline_pi_lower', 'pneumonia_influenza_covid','excess_pneumonia_influenza_covid')

res.ac2 <- res.ac1
names(res.ac2) <-
  c('year','week','day','week_end_date','baseline_all_cause','baseline_all_cause_upper', 'baseline_all-cause_lower', 'all_cause_deaths','excess_all_cause_deaths')

comb1 <- merge(res.ac2, res.pi2,
               by=c('year','week','day','week_end_date'))

comb2 <- merge(comb1,
               natl.all.data2[,c('week_end','covid19.nchs')],by.x='week_end_date', by.y='week_end')

write.csv(comb2,'./outputs/national_summary.csv' )
```


```{r pi_analysis_state, include=F}
#Run analysis
excess_deaths1.adjusted <-
  excessCases(ds = pi.data.clean2[!(pi.data.clean2$state %in% exclude.states),],
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "flu_pct_adj_lag1",
              #covs=c("flu_pct_adj", "flu_pct_adj_lag1", "flu_pct_adj_lag2"),
              use.syndromes = c("total_pni"),
              extrapolation.date = '2020-01-26',
              time.res='week')
```


```{r, echo=F}
 #dashboardPlot(excess_deaths1.adjusted)
```


```{r}
### Extract the quantities of interest

#Which syndrome do you want to plot, and over what time range?

syndrome.select <- 'total_pni' #which syndrome do you want to plot?
n.days<-52  #How many days to plot?
ds <- excess_deaths1.adjusted

#Extract the data needed to plot from the results

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred<-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr <-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

excess_deaths <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

excess_deaths_var <- excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred.var")

n.days <- 52
select.indices <- (length(dates1)-n.days):length(dates1)
dates<-dates1[select.indices]
states <- dimnames(pred)[[2]]
ages <- dimnames(pred)[[3]]
```

```{r summary_table_pi_state}
lpi.long <- melt(lpi[,,1])
names(lpi.long) <- c('index','state','lpi')

upi.long <- melt(upi[,,1])
names(upi.long) <- c('index','state','upi')

pred.long <- melt(lpi[,,1])
names(pred.long) <- c('index','state','pred')

obs.long <- melt(obs[,,1])
names(obs.long) <- c('index','state','obs')

excess_deaths.long <-
  melt(excess_deaths[,,1])


dates1.long <- rep(dates1, times=ncol(lpi))

mmwr.dates <- MMWRweek(dates1.long)
names(mmwr.dates) <- c('year', 'week', 'day')
pi.summary.state <-
  cbind.data.frame(mmwr.dates,
                   'week_end_date'=dates1.long+6,
                   'state'=pred.long[,2],
                   'baseline_pi'=pred.long[,3],
                   "baseline_pi_lower"=lpi.long[,3], "baseline_pi_upper"=upi.long[,3],
                   "pi_deaths"=obs.long[,3], 
                   "excess_pi_deaths"=excess_deaths.long[,3]
                                    )

```


## Figure 1: Observed weekly death rate vs seasonal baseline (+/-95% Prediction Interval) 
 The black line shows the observed proportion of deaths that were due to Pneumonia & Influenza (P&I) per week. The red line and shaded area represent the 95% Prediction Interval. The latest P&I data is for the week ending `r  max(dates1)+6`.



```{r fig1, fig.width=7.3, fig.height=6.9}

  par(mfrow=c(5,4), mar=c(2,4,2,1))
#states.cdc.order <- c(states.cdc[states.cdc %in% state.abb],
#                     states.cdc[!(states.cdc %in% state.abb)]) 
#plot.state.indices <- match(states.cdc.order,dimnames(pred)[[2]]  )

plot.state.rank <- cbind.data.frame(state.index=1:dim(rr)[2],state.rank= rank(-rr[dim(rr)[1],,1]))
plot.state.rank <- plot.state.rank[order(plot.state.rank$state.rank),]
plot.state.indices <- plot.state.rank$state.index

for(i in plot.state.indices){
  for(j in 1:dim(pred)[3]){
    y.range1<-range(c( pred[select.indices,,]/denom[select.indices,,],obs[select.indices,,]/denom[select.indices,,], upi[select.indices,,]/denom[select.indices,,]))
    if(states[i] %in% state.abb ){
      state.name.plot <-    
      state.name[match(states[i],state.abb)]
    }else{
    state.name.plot <- states[i]
  }
  plot(dates+6,
       pred[select.indices,i,j]/denom[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       xlab='',
       ylab='Proportion',
       #main=paste(states[i])
       )
  
  text(x=as.Date('2019-03-01'),y=y.range1[2], state.name.plot, pos=4,cex=0.9)

  points(dates+6,
         obs[select.indices,i,j]/denom[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates+6,
            rev(dates+6)),
          c(lpi[select.indices,i,j]/denom[select.indices,i,j],
            rev(upi[select.indices,i,j]/denom[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

## Figure 2: Observed deaths/expected deaths by state
Excess Deaths due to pneumonia and influenza

```{r, echo=F}
rr2<-rr[,,1]
date.mmwrdates <- mmwr_week(dates1)
mmwr.epiyr<- date.mmwrdates$mmwr_year
mmwr.epiyr[date.mmwrdates$mmwr_week<=26] <- mmwr.epiyr[date.mmwrdates$mmwr_week<=26] - 1

mmwr.epiwk <- date.mmwrdates$mmwr_week
mmwr.epiwk[date.mmwrdates$mmwr_week>=27]<-date.mmwrdates$mmwr_week[date.mmwrdates$mmwr_week>=27] - 52
mmwr.epiwk <- mmwr.epiwk +26
check<-cbind.data.frame(date.mmwrdates,mmwr.epiwk, mmwr.epiyr)
```

These plots show the Observed/Expected number of deaths due to pneumonia and influenza in each week for the 2019-20 year (red) compared to previous years (gray). Values close to 1 indicate that the values for that week are close to what would be expected based on the time of year and influenza activity.

```{r fig2, fig.width=7.3, fig.height=6.9}

  par(mfrow=c(5,4), mar=c(2,2,2,1), oma = c(3,3,0,0) + 0.1)
# states.cdc.order <- c(states.cdc[states.cdc %in% state.abb],
#                       states.cdc[!(states.cdc %in% state.abb)]) 
# plot.state.indices <- match( states.cdc.order,dimnames(pred)[[2]]  )

#plot.state.indices <- which(dimnames(pred)[[2]] %in% plot.states )
plot.state.rank <- cbind.data.frame(state.index=1:dim(rr)[2],state.rank= rank(-rr[dim(rr)[1],,1]))
plot.state.rank <- plot.state.rank[order(plot.state.rank$state.rank),]
plot.state.indices <- plot.state.rank$state.index
for(i in plot.state.indices){
    y.range1<-c(0.5,max(exp(rr2), na.rm=T))
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    
    if(states[i] %in% state.abb ){
      state.name.plot <-    
      state.name[match(states[i],state.abb)]
    }else{
    state.name.plot <- states[i]
    }
    
    matplot(ds2.c$epiwk         ,
       exp(ds2.c[,-1]),
       type='l',
       col=cols1,
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='', xlab='')
    abline(h=1, col='black')
     text(x=ds2.c$epiwk[1]*1.0 , y=y.range1[2]*0.95 , state.name.plot, pos=4, cex=0.9)
}
title(xlab = "Weeks Since start of July",
      ylab = "Observed/Expected",
      outer = TRUE, line = 1, cex=1.0)
```


```{r, eval=F}
## Estimated cumulative excess deaths across the available states, by week
## excess deaths

excess_deaths2 <- excess_deaths[dates1 >= as.Date(count.start.date),,1]
excess_deaths.state <- apply(excess_deaths2,2,sum)
cumsum_excess_deaths_state <- apply(excess_deaths2,2,cumsum)
#matplot(cumsum_excess_deaths_state, type='l', bty='l')

cumsum_excess_deaths_national <- cbind.data.frame('Week Ending'=dates1[dates1 >= as.Date(count.start.date)]+6, 'Excess P&I deaths'=apply(cumsum_excess_deaths_state,1, sum))
htmlTable(cumsum_excess_deaths_national)
```


## Figure 3: Reported number of COVID-19 deaths,compared with the excess deaths due to pneumonia and influenza in each week, by state. 
The red line shows the number of excess P&I cases +/-95% prediction intervals. The blue solid line shows the reported number of COVID-19 deaths for the same week (as compiled by the Johns Hopkins CSEE), and the dotted blue line shows the reported COVID-19 deaths for weeks in which the CDC data were not yet reliable. The grey dashed line represents number of tests performed per-capita in that week. Excess P&I estimates from recent weeks are adjusted for incomplete records using the "percent_complete" variable.

```{r}
excess_deaths.df <- cbind.data.frame('date'=dates1,excess_deaths[,,1])
excess_deaths.m <- melt(excess_deaths.df, id.vars =c('date'))
names(excess_deaths.m) <-c('date','state','excess_pi')

#excess_deaths_var
excess_deaths.var.df <- cbind.data.frame('date'=dates1,excess_deaths_var[,,1])
excess_deaths.var.m <- melt(excess_deaths.var.df, id.vars =c('date'))
names(excess_deaths.var.m) <-c('date','state','excess_pi_var')

excess_deaths.upi.df <- cbind.data.frame('date'=dates1,  (obs[,,1]-lpi[,,1] ))
excess_deaths.upi.m <- melt(excess_deaths.upi.df, id.vars =c('date'))
names(excess_deaths.upi.m) <-c('date','state','excess_deaths.upi')

excess_deaths.lpi.df <- cbind.data.frame('date'=dates1,(obs[,,1]-upi[,,1]) )
excess_deaths.lpi.m <- melt(excess_deaths.lpi.df, id.vars =c('date'))
names(excess_deaths.lpi.m) <-c('date','state','excess_deaths.lpi')

total_pi_df <- cbind.data.frame('date'=dates1,(obs[,,1]) )
total_pi_df.m <- melt(total_pi_df, id.vars =c('date'))
names(total_pi_df.m) <-c('date','state','total_pi')

#states.cdc <- unique(excess_deaths.m$state)
jh3 <- merge(excess_deaths.m, excess_deaths.lpi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.upi.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, total_pi_df.m, by=c('state','date') , all=T)
jh3 <- merge(jh3, excess_deaths.var.m, by=c('state','date') , all=T)

#jh3 <-jh3[jh3$state %in% states.cdc,]


jh3 <- merge( jh3, pi.data.clean2[c('state', "week_start",'percent_complete')], by.x=c('state','date') , by.y=c('state',"week_start"), all=T)

jh3 <- merge(jh3, test.ds3.spl, by=c('state','date'), all=T)
jh3$covid.track.death[is.na(jh3$covid.track.death)] <- 0


#Only plot out to 1 week ahead of the US P&I data
jh3 <- jh3[ jh3$date <= max(pi.data$"week_start" +weeks(1)),]
jh3 <- jh3[jh3$state %in% states.cdc,]
#jh3 <-jh3[!is.na(jh3$excess_pi ),]


jh3<-jh3[jh3$date >=as.Date(count.start.date),]
```

```{r fig3, fig.width=7.3, fig.height=6.9}
## uses covidtracking.com data for reported deaths

  par(mfrow=c(5,4), mar=c(2,4,3,1), oma = c(3,3,0,3) + 0.1)
ts.plot.func(ds.plot=jh3, death.var='covid.track.death' ,states.plot=unique(jh3$state),ylim.adj=2.0 )

title(xlab = "Date",
      ylab = "N Deaths",
      outer = TRUE, line = 1, cex=1.0)
   mtext(text="Tests/1000 people", line=2,side=4,outer=TRUE, cex=0.75, col='gray')

```

## Table 1: Excess P&I deaths and reported deaths due to COVID-19

```{r, table1}
count.excess <-jh3[jh3$date >= as.Date(count.start.date) & jh3$date <=max(pi.data$week_end) ,]

count.excess$excess_pi_adj <-count.excess$excess_pi/count.excess$percent_complete

count.excess$total_pi_adj <-count.excess$total_pi/count.excess$percent_complete

count.excess$excess_pi_var_ADJ <- count.excess$excess_pi_var/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi_adj','total_pi_adj','covid.track.death','excess_pi_var_ADJ')], by=list('state'=count.excess$state) ,FUN=sum)

count.excess2 <- merge(count.excess2, pop2, by='state')

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var_ADJ))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var_ADJ))

count.excess2$exces_ci_ucl_INC <- (count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var_ADJ))/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_ci_lcl_INC <- (count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var_ADJ))/count.excess2$census_bureau_pop_2019*1000000

count.excess2$exces_pi_INC <- count.excess2$excess_pi_adj/count.excess2$census_bureau_pop_2019*1000000

count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi_adj), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 
count.excess2$excess_pi_ci_INC <- paste0(round(count.excess2$exces_pi_INC,1), ' (',  round(count.excess2$exces_ci_lcl_INC,1), ', ', round(count.excess2$exces_ci_ucl_INC,1),')'  ) 

#sort by incidence
count.excess2 <- count.excess2[rev(order(count.excess2$exces_pi_INC)),]

count.excess2$total_pi_adj <- round(count.excess2$total_pi_adj)
count.excess3 <- count.excess2[, c('state','total_pi_adj','excess_pi_ci' ,'covid.track.death','excess_pi_ci_INC')]
names(count.excess3) <-c('State','Total P&I deaths', 'Excess P&I deaths', 'Reported COVID-19 deaths','Excess P&I deaths/1,000,000 people')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")

htmlTable(count.excess3, caption=paste0('Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from January 26, 2020 through ', last.date.format), rnames=F)
```


```{r}
## And finally,  use the NCHS COVID death data
#(Not yet available with state-level resolution)
#cdc.covid.url<- "https://data.cdc.gov/resource/hc4f-j6nb.json"
```


## Figure 4: Compare Excess P&I mortality vs Excess ILI

Here we compare the observed vs expected number of deaths due to pneumonia and influenza in each week compare to the observed vs expected number of outpatient visits for influenza-like illness (ILI) in each week. we would expect ILI (blue line) to increase earlier than deaths (red line)

```{r, include=F}
#ARCHIVE
ili.data <- runIfExpired('ilinet_state', ~ilinet(region = c("state")))
ili.data$state <- state.abb[match(ili.data$region, state.name)]
ili.data       <- ili.data[, c("state", "week_start", "ilitotal", "total_patients")]
ili.data       <- ili.data[!is.na(ili.data$total_patients),]
ili.data.spl   <- split(ili.data, ili.data$state)

min<-sapply(ili.data.spl, function(x)  min(x$total_patients))

state.select<-names(min)[which(min>0) ]
ili.data <- ili.data[ili.data$state %in% state.select,]
## Run the main analysis function, adjusting for flu using NREVSS data

ili.data2 <- merge(ili.data, hhs_states.cw, by='state')
ili.data2.agg <- aggregate(ili.data2[,c('ilitotal','total_patients')], by=list('state'=ili.data2$state_region,"week_start"=ili.data2$week_start), FUN=sum)

ili.data2.agg <- merge(ili.data2.agg, clin4.lags, by.x=c('state','week_start'), by.y=c('state', 'wk_date'))

excess_cases1 <-
  excessCases(ds = ili.data2.agg,
              datevar       = "week_start", 
              statevar      = "state",
              denom.var     = "total_patients",
              adj.flu       = "flu_pct_adj",
              use.syndromes = c("ilitotal"),
              extrapolation.date = "2020-03-01",
              time.res='week')

dates.ili <-
  excess_cases1[[1]][[1]][[1]]$date
  
rr.ili <-  excessExtract(ds = excess_cases1,
                syndrome = "ilitotal",
                extract.quantity = "resid1")
rr2.ili <- rr.ili[,,1]

date.mmwrdates.ili <- mmwr_week(dates.ili)
mmwr.epiyr.ili<- date.mmwrdates.ili$mmwr_year
mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] <- mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] - 1

mmwr.epiwk.ili <- date.mmwrdates.ili$mmwr_week
mmwr.epiwk.ili[date.mmwrdates.ili$mmwr_week>=27]<-date.mmwrdates.ili$mmwr_week[date.mmwrdates.ili$mmwr_week>=27] - 52
mmwr.epiwk.ili <- mmwr.epiwk.ili +26
```

```{r fig4, fig.width=7.3, fig.height=7}
common.states <- intersect(colnames(rr2), colnames(rr2.ili) )

# states.cdc.order <- c(common.states[common.states %in% state.abb],
#                       common.states[!(common.states %in% state.abb)]) 
# plot.state.indices <- match( states.cdc.order,common.states  )
rr.common.states <- rr[,common.states,1]
rr2.comp <-rr.common.states
rr2.ili.comp <- rr.ili[,common.states,]
plot.state.rank <- cbind.data.frame(state.index=1:dim(rr.common.states)[2],state.rank= rank(-rr.common.states[dim(rr.common.states)[1],]))
  plot.state.rank <- plot.state.rank[order(plot.state.rank$state.rank),]
  plot.state.indices <- plot.state.rank$state.index
  
date.start <-unique(pi.data$week_start[pi.data$weeknumber==27 & year(pi.data$week_start)==2019])
dates.plot4 <- seq.Date(from=date.start, length.out=52, by='week')

  par(mfrow=c(5,4), mar=c(2,4,2,2))

for(i in plot.state.indices){
    y.range1<-c(0,5)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2.comp[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    
    if(dimnames(rr.common.states)[[2]][i] %in% state.abb ){
      state.name.plot <-    
      state.name[match(dimnames(rr.common.states)[[2]][i],state.abb)]
    }else{
    state.name.plot <- dimnames(rr.common.states)[[2]][i]
    }
    
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    plot(dates.plot4         ,
       exp(ds2.c[,'2019']),
       type='l',
       col='#e41a1c',
       ylim=y.range1,
       xlim=c(as.Date(c('2019-07-01')),max(pi.data$week_end)),
       bty='l',
       lty=1,
       xlab='',
       xaxt='n',
       ylab='Observed/Expected'
       #main=common.states[i]
       )
    axis(1, at=as.Date(c('2019-07-01', '2019-10-01','2020-01-01', '2020-04-01', '2020-07-01')) , labels= c('Jul','Oct','Jan','Apr','Jul')  )

  es2<-cbind.data.frame('epiwk'=mmwr.epiwk.ili,'epiyr'=mmwr.epiyr.ili, rr=rr2.ili.comp[,i])
    es2.c<-dcast(es2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
     points(dates.plot4         ,
       exp(es2.c[,'2019']), type='l', col='#377eb8')
    abline(h=1, col='gray', lty=2)
    
    text(min(dates.plot4), y=0.2, state.name.plot,pos=4, cex=0.9, offset=0.1, xpd=NA)
    
 #   abline(v=as.Date(c('2020-03-07', '2020-03-21')), col=rgb(0,0,0,alpha=0.1))
}

```

```{r, eval=F}

#CHANGEPOINTS for excess ILI and excess P&I
cp.ds <- function(ds){
  spl<-matrix(NA, ncol=length(ds), nrow=length(ds))
  #Create the splines
for(i in 1:length(ds)){
  spl[,i]<- (1:length(ds)) -i
  spl[,i][spl[,i]<0]<-0
}
  aics<- rep(NA, ncol(spl))
  for(i in 1:ncol(spl)){
  mod1<- glm(ds ~ spl[,i] )
  aics[i]<-AIC(mod1)
  cp.date.index <- which(min(aics)==aics)
  }
  best.aic<-min(aics)
  w.aic<- exp(-0.5*(aics-best.aic))/sum(exp(-0.5*(aics-best.aic)))
  results <- list('w.aics'=w.aic,'cp.date.index'=cp.date.index )
  return(results)
} 
cp.ili <- apply(rr2.ili.comp[225:nrow(rr2.ili.comp),],2,cp.ds)
cp.ili.waics <- round(sapply(cp.ili,'[[', 'w.aics'),2)
cp.pi  <- apply(rr2.comp[225:nrow(rr2.comp),],2,cp.ds)
cp.pi.waics <- round(sapply(cp.pi,'[[', 'w.aics'),2)

#NOTE:: INDEX FOR ILI AND INDEX FOR PI are different--more week for ILI..need to convert back to DATE
#plot(cp.ili, cp.pi)

```



## Table S1: Excess deaths by week across the US, compared with provisional death counts from NCHS
Note excess P&I deaths and covidtracking.com numbers do not include Connecticut, North Carolina, and West Virginia. Combining together the exces P&I and COIVD-19 deaths without pneumonia diagnoses gives an estimate for number of deaths   
```{r}
count.excess <-jh3[jh3$date >= as.Date(count.start.date) & jh3$date <= max(pi.data$week_end) ,]

count.excess$excess_pi_adj <- count.excess$excess_pi/count.excess$percent_complete

count.excess$total_pi_adj <- count.excess$total_pi/count.excess$percent_complete

count.excess$excess_pi_var_ADJ <- count.excess$excess_pi_var/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi_adj','total_pi_adj','covid.track.death','excess_pi_var_ADJ')], by=list('date'=count.excess$date) ,FUN=sum)

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var_ADJ))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var_ADJ))


count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi_adj), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 

#sort by incidence
count.excess2 <- count.excess2[(order(count.excess2$date)),]

count.excess2$total_pi_adj <- round(count.excess2$total_pi_adj)
count.excess2$wk.end <-count.excess2$date+6
count.excess3 <- count.excess2[, c('wk.end','total_pi_adj','excess_pi_ci' ,'covid.track.death')]
names(count.excess3) <-c('Week ending','Total P&I deaths', 'Excess P&I deaths', 'All reported COVID-19 deaths (covidtracking.com)')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")

count.excess4 <- merge(count.excess3, cdc.summary.wk, by.x='Week ending', by.y='wk_end')

count.excess4$`NCHS Reported COVID-19 Deaths, no pneumonia code` <- 
  as.numeric(as.character(count.excess4$`NCHS Reported COVID-19 Deaths`)) -
  as.numeric(as.character(count.excess4$`NCHS Reported COVID-19 Deaths, with pneumonia code`))

count.excess4 <- count.excess4[, c("Week ending","Total P&I deaths","Excess P&I deaths","NCHS Reported COVID-19 Deaths, with pneumonia code", "NCHS Reported COVID-19 Deaths, no pneumonia code","NCHS Reported COVID-19 Deaths" , "All reported COVID-19 deaths (covidtracking.com)"  )]

htmlTable(count.excess4, caption=paste0('Table S1: Comparison of data sources: Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from January 26, 2020 through ', last.date.format), rnames=F)
```

```{r}
#Aggregate across weeks
count.excess.all <- apply(count.excess[,c('excess_pi_adj','total_pi_adj','covid.track.death','excess_pi_var_ADJ')] ,2,sum)
count.excess.all <- as.data.frame(matrix(count.excess.all, nrow=1))
names(count.excess.all) <- c('excess_pi_adj','total_pi_adj','covid.track.death','excess_pi_var_ADJ')

count.excess.all$exces_ci_ucl <- round(count.excess.all$excess_pi_adj + 1.96*sqrt(count.excess.all$excess_pi_var_ADJ))

count.excess.all$exces_ci_lcl <- round(count.excess.all$excess_pi_adj - 1.96*sqrt(count.excess.all$excess_pi_var_ADJ))


count.excess.all$excess_pi_ci <- paste0(round(count.excess.all$excess_pi_adj), '(',  count.excess.all$exces_ci_lcl, ', ', count.excess.all$exces_ci_ucl,')'  ) 

count.excess.all$excess_pi_ci

count.excess.all$covid.track.death

sum(as.numeric(as.character(count.excess4$`NCHS Reported COVID-19 Deaths`)))
```


## Figure S1: Excess all-cause deaths

Now instead of modeling the proportion of deaths that were due to pneumonia and influenza, we will model all-caused deaths, using the proportion under-reporting as the denominator

```{r fig.s1.setup}

#Use NY data separated by NY city and NY state
pi.data.agg.ny <- aggregate(pi.data.ny.separate[,c('total_pni','all_deaths','pct.comp.wgt')], by=list('state'=pi.data.ny.separate$state_region, 'week_start'=pi.data.ny.separate$week_start, 'week_end'=pi.data.ny.separate$week_end), FUN=sum, na.rm=T)
pi.data.agg.ny$percent_complete <- pi.data.agg.ny$pct.comp.wgt/pi.data.agg.ny$all_deaths

spl1.ny<-split(pi.data.agg.ny, pi.data.agg.ny$state)
min.state.ny <- lapply(spl1.ny, function(x){ x$miss.x<-min(x$total_pni, na.rm=T)
return(x)
                })
pi.data.clean.ny <- do.call('rbind.data.frame',min.state.ny)
pi.data.clean.ny <- pi.data.clean.ny[!is.na(pi.data.clean.ny$miss.x),]

pi.data.clean.ny$state.merge <- pi.data.clean.ny$state
pi.data.clean.ny$state.merge[pi.data.clean.ny$state=='NYC'] <-'NY'
pi.data.clean.ny <- merge(pi.data.clean.ny,clin4, by.x=c('state.merge', 'week_start'), by.y=c('state', 'wk_date'))

#Run analysis
pi.data.clean.ny$one <- 1
pi.data.clean.ny$percent_complete.adj <-pi.data.clean.ny$percent_complete - 0.5 #subtract =0.5 because 0.5 is added in code

pi.data.clean.ny$one <- 1000

excess_deaths1.ac <-
  excessCases(ds = pi.data.clean.ny,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "one",
              adj.flu       = "flu_pct_adj",
              #covs=c("flu_pct_adj", "flu_pct_adj_lag1", "flu_pct_adj_lag2"),
              use.syndromes = c("all_deaths"),
              extrapolation.date = '2020-01-26',
              time.res='week')
```

```{r}
ds <- excess_deaths1.ac
syndrome.select <- 'all_deaths'

excess_deaths.ac <- excessExtract(ds = excess_deaths1.ac,
                syndrome = 'all_deaths',
                extract.quantity = "unexplained.cases")
dates1.ac <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs.ac <-
  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred.ac<-  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr.ac <-  excessExtract(ds = excess_deaths1.ac,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

```

```{r summary_table_state_ac}
lpi.ac.long <- melt(lpi.ac[,,1])
names(lpi.ac.long) <- c('index','state','lpi.ac')

upi.ac.long <- melt(upi.ac[,,1])
names(upi.ac.long) <- c('index','state','upi.ac')

pred.ac.long <- melt(lpi.ac[,,1])
names(pred.ac.long) <- c('index','state','pred.ac')

obs.ac.long <- melt(obs.ac[,,1])
names(obs.ac.long) <- c('index','state','obs.ac')

excess_deaths.ac.long <-
  melt(excess_deaths.ac[,,1])


dates1.ac.long <- rep(dates1.ac, times=ncol(lpi.ac))

mmwr.dates.ac <- MMWRweek(dates1.ac.long)
names(mmwr.dates.ac) <- c('year', 'week', 'day')
ac.summary.state <-
  cbind.data.frame(mmwr.dates.ac,
                   'week_end_date'=dates1.ac.long+6,
                   'state'=pred.ac.long[,2],
                   'baseline'=pred.ac.long[,3],
                   "baseline_all-cause_lower"=lpi.ac.long[,3], "baseline_all_cause_upper"=upi.ac.long[,3],
                   "all_cause_deaths"=obs.ac.long[,3], 
                   "excess_all_cause_deaths"=excess_deaths.ac.long[,3])

ac.summary.state$week_start_date <-
  ac.summary.state$week_end_date-6
                                                    
jh3a <- jh3
jh3a$week_start <- jh3a$date 
jh3a <- jh3a[,c('state','week_start',"covid.track.death","test.week.per.capita",'testN' )]

ac.summary.state2 <- merge(ac.summary.state, jh3a,by.x=c('state','week_start_date'), by.y=c('state', 'week_start'), all=T)

ac.summary.state2$week_start_date <- NULL

ac.summary.state3 <- 
  merge(ac.summary.state2,pi.summary.state ,
        by=c('year','week','day','state','week_end_date'), all=T)

write.csv(ac.summary.state3,'./outputs/state.summary.csv')

end.summary <- ac.summary.state3[ac.summary.state3$year==2020 & ac.summary.state3$week>=10 & ac.summary.state3$week<=14 ,]

agg.est <- 
  aggregate(end.summary[,c('excess_pi_deaths',"excess_all_cause_deaths","all_cause_deaths",'covid.track.death')], by=list('state'=end.summary$state), FUN=sum) 
agg.est$ratio <-
  agg.est$excess_all_cause_deaths/agg.est$covid.track.death
```

```{r}
# compare excess all-cause deaths with excess P&I deaths

dates.select <- which(dates1 >= count.start.date)
#plot( excess_deaths[dates.select,,1],excess_deaths.ac[dates.select,,1])

```


The black line shows the observed number of *all* deaths per week, regardless of cause. The red line and shaded area represent the 95% Prediction Interval. The latest data is for the week ending `r  max(dates1)+6`. Note that these are **adjusted** for percent completeness of the data. There are clear jumps in all-cause mortality in NY and NJ, other states less clear, probably due to reporting delays

```{r fig.s1, fig.width=8, fig.height=7}
  par(mfrow=c(5,4), mar=c(3,4,1,1))

plot.state.rank <- cbind.data.frame(state.index=1:dim(rr.ac)[2],state.rank= rank(-rr.ac[dim(rr.ac)[1],,1]))
plot.state.rank <- plot.state.rank[order(plot.state.rank$state.rank),]
plot.state.indices <- plot.state.rank$state.index
states.ac <-dimnames(pred.ac)[[2]]

select.indices <- which(dates1.ac >= as.Date('2009-04-01'))

for(i in plot.state.indices){
  for(j in 1:dim(pred.ac)[3]){
    y.range1<-range(c( pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j], upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j]))
  plot(dates1.ac,
       pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       ylab='Number',
       main=paste(states.ac[i])
  )

  points(dates1.ac,
         obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates1.ac,
            rev(dates1.ac)),
          c(lpi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
            rev(upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

```{r}
#Sum excess across all states
select.indices <-which(dates1.ac >=as.Date('2020-01-26'))
pred.all <- apply(pred.ac[select.indices,,1],1,sum)
obs.all <- apply(obs.ac[select.indices,,1],1,sum)
obs.all - pred.all
```


```{r, eval=F}
#Just NY
par(mfrow=c(1,2))
  plot.state.indices <- which(dimnames(pred.ac)[[2]] %in% c('NY','NYC') )
for(i in plot.state.indices){
  for(j in 1:dim(pred.ac)[3]){
    y.range1<-range(c( pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j], upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],0))
  plot(dates1.ac[select.indices],
       pred.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
       type='l',
       col='red',
       ylim=c(0,3000),
       bty='l',
       ylab='Number',
       main=paste(states.ac[i],ages[j]))

  points(dates1.ac[select.indices],
         obs.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates1.ac[select.indices],
            rev(dates1.ac[select.indices])),
          c(lpi.ac[select.indices,i,j]/denom.ac[select.indices,i,j],
            rev(upi.ac[select.indices,i,j]/denom.ac[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```


## Figure S2: Excess all-cause deaths for NY state (incl NYC) vs Excess deaths due to pneumonia and influenza

```{r fig.s2, fig.width=8, fig.height=3}
 dates.select<- which(dates1.ac>=as.Date('2020-02-09'))
 dates.select.pi<- which(dates1>=as.Date('2020-02-09'))
#
report.ny <- test.ds3.spl[test.ds3.spl$state=='NY' & test.ds3.spl$date %in% dates1.ac[dates.select], ]
report.ny <- report.ny[order(report.ny$date),]

excess_all_cause <- obs.ac - pred.ac

excess_pi <- obs - pred

par(mfrow=c(1,2), mar=c(2,4,1,1))

# plot(dates1.ac,excess_all_cause[,'NY',1]+excess_all_cause[,'NYC',1], type='l', bty='l', ylab='Excess deaths (N)', main='New York, incl NYC')
# points(dates1,excess_pi[,'NY',1], type='l', col='red')
# legend(as.Date('2014-01-01'), 1500, legend=c("Excess all-cause deaths", "Excess P&I deaths"),
#        col=c("black", "red"), lty=1, cex=0.8)
# 
# plot(dates1.ac[],excess_all_cause[,'NJ',1], type='l', bty='l', ylab='Excess deaths (N)', main='New Jersey')
# points(dates1,excess_pi[,'NJ',1], type='l', col='red')
# 
# sum(excess_all_cause[dates.select,'NYC',1] )

plot(dates1.ac[dates.select],excess_all_cause[dates.select,'NY',1]+excess_all_cause[dates.select,'NYC',1], type='l', bty='l', ylab='Excess deaths (N)', main='New York, incl NYC')
points(dates1[dates.select.pi],excess_pi[dates.select.pi,'NY',1], type='l', col='red')
points(report.ny$date, report.ny$covid.track.death, col='blue', type='l', lty=2)
 legend(as.Date('2020-02-09'), 1500, legend=c("Excess all-cause deaths",   "Excess P&I deaths", 'Reported COVID-19'),
       col=c("black", "red", 'blue'), lty=c(1,1,2), cex=0.8) 

plot(dates1.ac[dates.select],excess_all_cause[dates.select,'NJ',1], type='l', bty='l', ylab='Excess deaths (N)', main='New Jersey')
points(dates1[dates.select.pi],excess_pi[dates.select.pi,'NJ',1], type='l', col='red')

```


Fold difference between excess all-cause mortality and excess P&I mortality in NJ and NY. Shows that excess P&I accounts for ~50-60% of the excess deaths
```{r}
excess_all_cause[dim(excess_all_cause)[1],'NJ',1]/excess_pi[dim(excess_pi)[1],'NJ',1]

(excess_all_cause[dim(excess_all_cause)[1],'NY',1]+excess_all_cause[dim(excess_all_cause)[1],'NYC',1])/excess_pi[dim(excess_pi)[1],'NY',1] 

```
 
## Figure S3: NYC only data P&I vs all-cause excess deaths
```{r nyc}
nyc.data <- pi.data.clean.ny[pi.data.clean.ny$state=='NYC',]
nyc.data$one <- 1
nyc.pi <-
  excessCases(ds = nyc.data,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "none",
              use.syndromes = c("total_pni"),
              extrapolation.date = '2020-01-26',
              time.res='week')
nyc.acm <-
  excessCases(ds = nyc.data,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "one",
              adj.flu       = "none",
              use.syndromes = c("all_deaths"),
              extrapolation.date = '2020-01-26',
              time.res='week')

nyc.excess.pi <-  excessExtract(ds = 
                nyc.pi,
                syndrome = 'total_pni',
                extract.quantity = "unexplained.cases")
nyc.excess.acm <-  excessExtract(ds = 
                nyc.acm,
                syndrome = 'all_deaths',
                extract.quantity = "unexplained.cases")

dates.select<- which(nyc.data$week_start>=as.Date('2020-02-09'))
dates2 <- nyc.data$week_end
plot(dates2[dates.select],nyc.excess.acm[dates.select,1,1], type='l', bty='l', ylab='Excess deaths (N)', main='New York City')
points(dates2[dates.select],nyc.excess.pi[dates.select,1,1], type='l', col='red')
legend(as.Date('2016-01-01'), 300, legend=c("Excess all-cause deaths", "Excess P&I deaths"),
       col=c("black", "red"), lty=1, cex=0.8)

```
Ratio of all-cause vs P&I deaths in NYC. In NYC, only 25% of excess deaths were coded as P&I
```{r}
dates.select2 <- dates.select<- which(nyc.data$week_start>=as.Date('2020-03-15'))
 
nyc.excess.pi[dates.select2,1,1]/nyc.excess.acm[dates.select2,1,1]

sum(nyc.excess.acm[dates.select2,1,1])

```

## Table S2: Proportion of COVID-19 deaths with a pneumonia code, by state, through most recent date
Note these values will be greater than those in other tables, which are 2 weeks behind

```{r}
cdc.summary.state.clean <- cdc.summary.state[!is.nan(cdc.summary.state$covid.pct.pneum),c('ind', "NCHS Reported COVID-19 Deaths","NCHS Reported COVID-19 Deaths, with pneumonia code",'covid.pct.pneum')]

names(cdc.summary.state.clean) <- c('State',"NCHS Reported COVID-19 Deaths","NCHS Reported COVID-19 Deaths, with pneumonia code",'Proportion with pneumonia')
#cdc.summary.state.clean

cdc.summary.state.clean$`Proportion with pneumonia` <- round(cdc.summary.state.clean$`Proportion with pneumonia`,2)
htmlTable(cdc.summary.state.clean, rnames=F)
```


## Table S3: Do not adjust for flu

```{r, tables1}
count.excess <-jh3[jh3$date >= as.Date(count.start.date) & jh3$date <=max(pi.data$week_end) ,]

count.excess$excess_pi_adj <-count.excess$excess_pi/count.excess$percent_complete

count.excess2 <- aggregate(count.excess[,c('excess_pi','excess_pi_adj','covid.track.death','excess_pi_var')], by=list('state'=count.excess$state) ,FUN=sum)

count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi + 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi - 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_ucl_adj <- round(count.excess2$excess_pi_adj + 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$exces_ci_lcl_adj <- round(count.excess2$excess_pi_adj - 1.96*sqrt(count.excess2$excess_pi_var))

count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 

count.excess2$excess_pi_ci_adj <- paste0(round(count.excess2$excess_pi_adj ), '(',  count.excess2$exces_ci_lcl_adj , ', ', count.excess2$exces_ci_ucl_adj ,')'  ) 
count.excess2 <- count.excess2[rev(order(count.excess2$excess_pi)),]

count.excess3 <- count.excess2[, c('state','excess_pi_ci', 'excess_pi_ci_adj', 'covid.track.death')]
names(count.excess3) <-c('State', 'Excess P&I deaths (unadjusted)','Excess P&I deaths (adjusted)', 'Reported COVID-19 deaths')
last.date.format<-max(pi.data$week_end)
last.date.format<-format(last.date.format,
                         "%b %d, %Y")

htmlTable(count.excess3, caption=paste0('Table S3. Observed and Excess deaths due to pneumonia & influenza, and COVID-19, from January 26, 2020 through ', last.date.format), rnames=F)
```


## Table S4: do not adjust for delayed reporting

```{r, include=F} 
#Run analysis 
excess_deaths1.unadjusted <-
  excessCases(ds = pi.data.clean2,
              datevar       = "week_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "none",
              use.syndromes = c("total_pni"),
              extrapolation.date = "2020-03-01",
              time.res='week')

syndrome.select <- 'total_pni' #which syndrome do you want to plot?
n.days<-52  #How many days to plot?
ds.unadj <- excess_deaths1.unadjusted
```
 
```{r}
#Extract the data needed to plot from the results

dates1.unadj <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases.unadj <-
  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr.unadj <-
  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom.unadj <-
  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi.unadj <-
  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi.unadj <-
  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs.unadj <-
  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred.unadj<-  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr.unadj <-  excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

excess_deaths.unadj <- excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

excess_deaths_var.unadj <- excessExtract(ds = ds.unadj,
                syndrome = syndrome.select,
                extract.quantity = "pred.var")
```

```{r}
excess_deaths.df.unadj <- cbind.data.frame('date'=dates1,excess_deaths.unadj[,,1])
excess_deaths.m.unadj <- melt(excess_deaths.df.unadj, id.vars =c('date'))
names(excess_deaths.m.unadj) <-c('date','state','excess_pi')

#excess_deaths_var
excess_deaths.var.df.unadj <- cbind.data.frame('date'=dates1,excess_deaths_var.unadj[,,1])
excess_deaths.var.m.unadj <- melt(excess_deaths.var.df.unadj, id.vars =c('date'))
names(excess_deaths.var.m.unadj) <-c('date','state','excess_pi_var')

excess_deaths.upi.df.unadj <- cbind.data.frame('date'=dates1,  (obs[,,1]-lpi[,,1] ))
excess_deaths.upi.m.unadj <- melt(excess_deaths.upi.df.unadj, id.vars =c('date'))
names(excess_deaths.upi.m.unadj) <-c('date','state','excess_deaths.upi')

excess_deaths.lpi.df.unadj <- cbind.data.frame('date'=dates1,(obs[,,1]-upi[,,1]) )
excess_deaths.lpi.m.unadj <- melt(excess_deaths.lpi.df.unadj, id.vars =c('date'))
names(excess_deaths.lpi.m.unadj) <-c('date','state','excess_deaths.lpi')

total_pi_df.unadj <- cbind.data.frame('date'=dates1,(obs[,,1]) )
total_pi_df.m.unadj <- melt(total_pi_df.unadj, id.vars =c('date'))
names(total_pi_df.m.unadj) <-c('date','state','total_pi')

#states.cdc <- unique(excess_deaths.m$state)
jh3.unadj <- merge(excess_deaths.m.unadj, excess_deaths.lpi.m.unadj, by=c('state','date') , all=T)
jh3.unadj <- merge(jh3.unadj, excess_deaths.upi.m.unadj, by=c('state','date') , all=T)
jh3.unadj <- merge(jh3.unadj, total_pi_df.m.unadj, by=c('state','date') , all=T)
jh3.unadj <- merge(jh3.unadj, excess_deaths.var.m, by=c('state','date') , all=T)

#jh3 <-jh3[jh3$state %in% states.cdc,]


jh3.unadj <- merge( jh3.unadj, pi.data.clean2[c('state', "week_start",'percent_complete')], by.x=c('state','date') , by.y=c('state',"week_start"), all=T)

jh3.unadj <- merge(jh3.unadj, test.ds3.spl, by=c('state','date'), all=T)
jh3.unadj$covid.track.death[is.na(jh3$covid.track.death)] <- 0


#Only plot out to 1 week ahead of the US P&I data
jh3.unadj <- jh3.unadj[ jh3.unadj$date <= max(pi.data$"week_start" +weeks(1)),]
jh3.unadj <- jh3.unadj[jh3.unadj$state %in% states.cdc,]
#jh3 <-jh3[!is.na(jh3$excess_pi ),]


jh3.unadj<-jh3.unadj[jh3.unadj$date >=as.Date(count.start.date),]

```

```{r}
excess_table.func <- function(ds){
count.excess <- ds[ds$date >=count.start.date & ds$date <= max(pi.data$week_end), ]

      count.excess$excess_pi <-count.excess$excess_pi/count.excess$percent_complete
    
    count.excess$total_pi_adj <-count.excess$total_pi/count.excess$percent_complete
    
    count.excess2 <- aggregate(count.excess[,c('excess_pi','total_pi','covid.track.death','excess_pi_var')], by=list('state'=count.excess$state) ,FUN=sum)
    
    count.excess2 <- merge(count.excess2, pop2, by='state')
    
    count.excess2$exces_ci_ucl <- round(count.excess2$excess_pi + 1.96*sqrt(count.excess2$excess_pi_var))
    
    count.excess2$exces_ci_lcl <- round(count.excess2$excess_pi - 1.96*sqrt(count.excess2$excess_pi_var))
    
    count.excess2$exces_ci_ucl_INC <- (count.excess2$excess_pi + 1.96*sqrt(count.excess2$excess_pi_var))/count.excess2$census_bureau_pop_2019*1000000
    
    count.excess2$exces_ci_lcl_INC <- (count.excess2$excess_pi - 1.96*sqrt(count.excess2$excess_pi_var))/count.excess2$census_bureau_pop_2019*1000000
    
    count.excess2$exces_pi_INC <- count.excess2$excess_pi/count.excess2$census_bureau_pop_2019*1000000
    
    count.excess2$excess_pi_ci <- paste0(round(count.excess2$excess_pi), '(',  count.excess2$exces_ci_lcl, ', ', count.excess2$exces_ci_ucl,')'  ) 
    count.excess2$excess_pi_ci_INC <- paste0(round(count.excess2$exces_pi_INC,1), ' (',  round(count.excess2$exces_ci_lcl_INC,1), ', ', round(count.excess2$exces_ci_ucl_INC,1),')'  ) 
     return(count.excess2)
}

excess.adj <- excess_table.func(ds=jh3)
excess.unadj <- excess_table.func(ds=jh3.unadj)


compare.excess <- merge(excess.unadj[,c('state',"excess_pi_ci")], excess.adj[,c('state',"excess_pi_ci", 'excess_pi')], by='state')
names(compare.excess) <- c('state','Excess P&I (unadjusted)', 'Excess P&I (adjusted for reporting delay)', 'excess_pi_count')

compare.excess <- compare.excess[rev(order(compare.excess$excess_pi_count)),]

compare.excess$excess_pi_count <-NULL

htmlTable(compare.excess, caption=paste0('Excess deaths due to pneumonia & influenza from January 26, 2020 through ', last.date.format, ' with or without adjustment for delayed reporting'), rnames=F)
```



Simple fig
```{r, fig.width=3, fig.height=6}
agg1<-pi.data
agg1$weeknumber <- as.numeric(as.character(agg1$weeknumber))
agg1$year <- year(agg1$week_end)
last.week.num <- 
  unique(agg1$weeknumber[agg1$week_end == (last.date.obtained-params$n.days.filter)])

agg2 <- 
  agg1[(agg1$weeknumber >= 10 & agg1$weeknumber <=(last.week.num)),]

agg3 <- 
  aggregate(agg2[,'all_deaths' ], 
            by=list('state'=agg2$region_name,'year'=agg2$year ), FUN=sum)

agg3.m <- melt(agg3, id.vars=c('state', 'year'))
agg3.c <- dcast(agg3.m, year~state)

agg3.c.std <- apply(agg3.c, 2,function(x) (x-mean(x))/sd(x))
agg3.c.std.m <- melt(agg3.c.std, id.vars='year')
agg3.c.std.m$stateN <-
  as.numeric(as.factor(agg3.c.std.m$Var2))
agg3.c.std.m <-
  agg3.c.std.m[agg3.c.std.m$Var2 %in% c('New York','New York City', 'New Jersey', 'Pennsylvania', 'Michigan', 'Washington', 'Louisiana', 'Florida'),]
agg3.c.std.m$Var2 <- factor(agg3.c.std.m$Var2)
agg3.c.std.m$stateN <-
  as.numeric(as.factor(agg3.c.std.m$Var2))

cols <- c(rep(rgb(0,0,0,alpha=0.3),(max(agg3.c.std.m$Var1)-1) ),rgb(1,0,0, alpha=0.8))
par(mar=c(2,6,1,1))
plot(agg3.c.std.m$value,
     agg3.c.std.m$stateN + rnorm(nrow(agg3.c.std.m),0,0.05),
     bty='l' , col=cols, pch=16, yaxt='n', ylab='')
axis(side=2, at= 1:max(agg3.c.std.m$stateN),
     labels=unique(agg3.c.std.m$Var2), las=2)


```

